### 2018-04-30 paper/ms.tex
update input file to use 1arcsec match from M.Bedell

\documentclass[preprint2]{aastex62}

\bibliographystyle{aasjournal}
\usepackage{graphicx}
\usepackage[suffix=]{epstopdf}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{url}
\usepackage{xspace}

%    Make Scientific Notation
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}

% make the word Kepler italicized
\newcommand{\Kepler}{\textsl{Kepler}\xspace}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%
\title{Rotating Stars from \Kepler Observed with Gaia DR2}

\shorttitle{Rotating Stars}
\shortauthors{Davenport et al.}

\correspondingauthor{James. R. A. Davenport}
\email{James.Davenport@wwu.edu}

\author{James. R. A. Davenport}
\altaffiliation{NSF Astronomy and Astrophysics Postdoctoral Fellow}
\altaffiliation{DIRAC Fellow}
\affiliation{Department of Physics \& Astronomy, Western Washington University, 516 High St., Bellingham, WA 98225, USA}
\affiliation{Department of Astronomy, University of Washington, Seattle, WA 98195, USA}

\author{Kevin R. Covey}
\affiliation{Department of Physics \& Astronomy, Western Washington University, 516 High St., Bellingham, WA 98225, USA}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Astrometric data from the recent Gaia Data Release 2 has been matched against the sample of stars from \Kepler with known rotation periods. 
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

The \Kepler mission \citep{borucki2010}

\citet{mcquillan2013} bimodal period distribution 
\citep{mcquillan2014}.

\citet{davenport2017} looked at this with Gaia DR1, extending period gap to G dwarfs, pointing to SFH as cause. sub giant contamination was the thing. Now we can finally follow up on this and study the full \Kepler rotation period sample

 \citep{gaia} mission

In this paper

%%%%%%%%%%%%%%%%%%%%%%
\section{The \Kepler--Gaia Data}
%Rotation periods in this study come from \citet{mcquillan2014}, who performed an Auto-Correlation Function analysis of \Kepler stars cooler than 6500 K that had at least $\sim$2 years of observation. The periods recovered from this approach generally agree very well with those found via Lomb-Scargle Periodograms \citep[e.g.][]{reinhold2013,aigrain2015}. Sources with multiple distinct periods, such as from binary systems with two spotted stars \citep[e.g.][]{lurie2015} are detected by \citet{mcquillan2014}, but are not included in the following analysis.

%The Gaia Data Release 1 (DR1) provides astrometric positions for over $10^9$ sources from the first year of observation with Gaia. The Tycho-Gaia Astrometric Solution (TGAS) measures improved proper motions and parallaxes for 2 million nearby, bright sources by extending the astrometric solutions from Tycho and Hipparcos. While the TGAS data are not a complete astrometric survey, and have possible systematics in the reported parallaxes \citep{gaia_dr1,stassun2016}, they represent a significant improvement in the astrometry and kinematics available for stars in the \Kepler field.

%%%%%%%%%%%%%%%%%%%%%%
\section{Selecting Main Sequence Stars}

\begin{figure*}[]
\centering
\includegraphics[width=6in]{../figures/cmd}
\caption{
color--magnitude diagram for 30,928 \Kepler stars included in Gaia DR2, colored by the rotation period measured from \citet{mcquillan2014}.}
\label{fig:cmd}
\end{figure*}

%Though \citet{mcquillan2014} attempted to only measure periods for dwarf stars, the sample of \Kepler--Gaia matched stars contains both main sequence dwarfs and evolved stars (giants and subgiants). Previous studies have shown  significant contamination by giants or subgiants can affect the implied variability properties of dwarf stars \citep{ciardi2011,mann2012}. Therefore to properly understand the nature of the period distribution and its implications for age-dating field stars, a robust sample of main sequence stars must be selected.
%
%Faint stars were removed by requiring sources have $G$-band flux errors $<1$\%. To ensure accurate distances, and therefore luminosities, parallaxes were required to have errors $<0.4$ mas. These cuts left a total of 894 stars from the \Kepler--TGAS matched sample (68\%). The Hertzsprung--Russell (HR) diagram for these stars is shown in Figure \ref{fig:HR}, with each point colored by its \Kepler-measured rotation period. Example isochrones from the \citet{bressan2012} grid are shown for two ages. A systematic offset of $\sim$0.5 magnitudes is found between the measured absolute $G$-band and the isochrone's main sequence. This offset is likely due to calibration differences between the nominal and actual $G$-band (A. Brown, private communication).
%
%
%The HR-period diagram in Figure \ref{fig:HR} shows stars with a range of evolutionary states, and could help test post-main sequence angular momentum evolution models \citep[e.g.][]{donascimento2012}. Outliers in this diagram are either due to erroneous cross-matching in the \Kepler and Gaia catalogs, or represent interesting systems such as rare binary star configurations or stars that have undergone mergers or ingested giant planets \citep{massarotti2008,tayar2015}. For example, examining the 2 Micron All Sky Survey \citep[2MASS][]{2mass} image on SIMBAD for 
%the rapidly rotating star at $T_{eff}=4500$ K, $M_G=4.5$ mag, $P_{rot}$=0.652 d, in Figure \ref{fig:HR} (KIC 07957709), this source appears highly contaminated with three point sources clustered within $\sim$10 arcseconds, likely leading to an erroneous position in the HR diagram, and possibly incorrect variability measurements with \Kepler.
%Investigating all such outliers in Figure \ref{fig:HR} is beyond the scope of this work, but these targets are worth further study as they may reveal new physics.
%
%
%
%Main sequence stars were selected using a simple cut around 300 Myr isochrone. Given the systematic offset of $\sim$0.5 mag between the Myr isochrone and the observed $M_G$ values, a fairly wide band of stars ($0\le \Delta M_G \le 1$) was selected as being ``close to the main sequence''. The final sample included 440 stars. This simplistic cut is not a robust dwarf--giant, nor single--binary star separator, but serves to select a sample of mostly main sequence stars for the illustrative purpose of this work. More precise selection will require an improved isochrone track, as well as updated parallaxes from the full Gaia DR2.
%

%%%%%%%%%%%%%%%%%%%%%%%
%\section{Extending the Spin-Down Gap}
%A bimodal period distribution was first discovered by \citet{mcquillan2013} for \Kepler M dwarfs, who found a dearth of objects with periods around $\sim$25 days. Follow-up work by \citet{mcquillan2014} found this bimodality extended to K dwarfs, up to $T_{eff}\sim5500$. Figure \ref{fig:gyro} shows the rotation period distribution for the final 440 star sample of likely main sequence stars. The bimodality appears to extend smoothly through to the hottest stars in this sample. 
%
%While these periods for field stars were robustly measured by \citet{mcquillan2014} and others, the bimodality did not appear in previous \Kepler work due to the high contamination rate by subgiants for these bluer, hotter stars.  414 rotating stars had acceptable photometric and parallax uncertainties in TGAS, but were culled from this sample for having $M_G$ luminosities higher than the main sequence cut in \S3 above. Note that distributions of $\log g$ values from the Kepler Input Catalog \citep{brown2011a} for both the main sequence and subgiant stars were not statistically different.

\begin{figure*}[]
\centering
\includegraphics[width=3.5in]{../figures/rot_dist_0}
\includegraphics[width=3.5in]{../figures/rot_dist_350}
\includegraphics[width=3.5in]{../figures/rot_dist_525}
\includegraphics[width=3.5in]{../figures/rot_dist_600}
\includegraphics[width=3.5in]{../figures/rot_dist_900}
\caption{
}
%\label{fig:cmd}
\end{figure*}

\begin{figure}[]
\centering
\includegraphics[width=3.5in]{../figures/B_V_rot}
\caption{
}
%\label{fig:cmd}
\end{figure}

\begin{figure}[]
\centering
\includegraphics[width=3.5in]{../figures/delta_per}
\caption{
}
%\label{fig:cmd}
\end{figure}

\begin{figure*}[]
\centering
\includegraphics[width=6.5in]{../figures/cmd_zoom}
\caption{
}
%\label{fig:cmd}
\end{figure*}

%\begin{figure}[]
%\centering
%\includegraphics[width=3.5in]{../figures/iso_age}
%\caption{
%}
%%\label{fig:cmd}
%\end{figure}

%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%
%Using a combination of data from \Kepler and Gaia DR1, I have explored the rotation period distribution for 440 nearby main sequence stars. A bimodal rotation period distribution has been found in stars with temperatures ranging from 5000 K to 6500 K. This feature matches that found in cooler stars from \Kepler, but was only revealed thanks to the enhanced ability to distinguish dwarfs from subgiants using Gaia data. 
%A tenuous difference in the TGAS total proper motion for stars in the fast and slow rotating groups is found, which is in agreement with the findings for cool stars by \citet{mcquillan2013}.
%
%While a definitive explanation for this period bimodality has not been reached, the findings to date seem to favor stellar ages as the cause. In this scenario the star formation history for nearby stars would be dominated by two epochs of star formation, one short event centered at a few hundred Myr, and one long event centered at a few Gyr (slightly younger than the Sun). It is also worth noting that the space volume probed by the TGAS sample investigated here is very similar to that covered by the temperature-selected cool star sample in \citet{mcquillan2013}. The median parallax distance for stars in this work is 285 pc, while the median isochrone distance for the K and M dwarfs is $\sim$216 pc. This points to the period distribution being a localized age artifact. 
%Determining how localized this age distribution is, and if it can be confirmed for stars across the HR diagram including giants, is a key goal for future Gaia data releases.
%
%
%The period bimodality may yet be a manifestation of the ``Vaughan-Preston'' gap observed in chromospheric activity indicators from solar type stars. Such a feature has also been discussed for rotating stars by \citet{kado-fong2016}. Given that the mass range for the bimodality explored here and in \citet{mcquillan2014} covers stars with solar-type dynamos (those having a tachocline, late F through early M) such a model cannot be fully ruled out at this time. Though there have been many rotation studies for cool stars \citep[e.g.][]{irwin2011,newton2016,stelzer2016} too few rotation periods have been measured for stars across the ``fully convective boundary'' ($T_{eff}<3000$ K, spectral type $\sim$M4) to tell if the bimodal period feature continues to cooler temperatures, which would support the age distribution model. 
%If the bimodality is due to stars crossing a phase of rapid angular momentum evolution, we would expect to see it in stellar clusters at or near the critical age. The lack of this feature in the clusters observed to date could be due to no cluster being close enough to the critical age, which the gyrochrone in Figure \ref{fig:gyro} shows is near 600 Myr. Further studies of rotation periods for stars in intermediate age open clusters (e.g. the Hyades) may help solve this mystery \citep[e.g.][]{douglas2014}.
%
%Finally, this exploratory work has highlighted the utility of using astrometric data from Gaia combined with detailed light curve statistics from \Kepler to reveal hidden substructure in the properties of field stars. Looking forward to the astrometric precision of future Gaia data releases, this combination will be effective at separating dwarf stars from subgiants for nearly the entire \Kepler and K2 databases, and enable accurate age maps for field stars.

%%%%%%%%%%%%%%%%%
\acknowledgments

JRAD is supported by an NSF Astronomy and Astrophysics Postdoctoral Fellowship under award AST-1501418.

This work made use of the \url{gaia-kepler.fun} crossmatch database created by Megan Bedell.

This project was developed as part of the 2018 NYC Gaia Sprint, hosted by the Center for Computational Astrophysics at the Simons Foundation in New York City.

%This research made use of the cross-match service, and the SIMBAD database, provided by CDS, Strasbourg, France.
%
%This work has made use of data from the European Space Agency (ESA)
%mission {\it Gaia} (\url{http://www.cosmos.esa.int/gaia}), processed by
%the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
%\url{http://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
%for the DPAC has been provided by national institutions, in particular
%the institutions participating in the {\it Gaia} Multilateral Agreement.

\bibliography{/Users/james/Dropbox/references}

\end{document}

### 2018-04-30 paper/ms.tex
outlining things

@@
 While the TGAS data are not a complete astrometric survey, and have possible systematics in the reported parallaxes \citep{gaia_dr1,stassun2016}, they represent a significant improvement in the astrometry and kinematics available for stars in the \Kepler field.
+
+data xmatched by M. Bedell between Kepler DR25 (CITE) and Gaia DR2 (CITE), using 1arcsec radius.
+distances using the updated prescription from \citet{bailer-jones2018}
 
@@
 \includegraphics[width=3.5in]{../figures/B_V_rot}
-\caption{
+\caption{converted to B-V using Padova isochrone colors. 3 gyrochrones from \citet{meibom2009} with ages 300 Myr (dark red), 600 Myr (bright red), and 1.2 Gyr (blue) are shown.
 }

### 2018-05-05 paper/ms.tex
update plots

--- 2018-04-30 paper/ms.tex
@@
 
-In this paper
+In this letter
 
@@
 
+data xmatched by M. Bedell between Kepler DR25 (CITE) and Gaia DR2 (CITE), using 1arcsec radius.
+distances using the updated prescription from \citet{bailer-jones2018}
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 \caption{
-color--magnitude diagram for 30,928 \Kepler stars included in Gaia DR2, colored by the rotation period measured from \citet{mcquillan2014}.}
+Color--magnitude diagram for 30,305 \Kepler stars from the \citet{mcquillan2014} sample that are included in Gaia DR2, colored by their measured rotation period.
+For reference we show a $10^9$ year MIST isochrone (black line) used to select likely main sequence, single stars (black line).
+A track of binary  stars is apparent $\sim$0.7 mag above the main sequence.
+As in \citet{davenport2017}, we find significant contamination of the rotation period sample for bluer stars by sub-giants.}
 \label{fig:cmd}
@@
 \includegraphics[width=3.5in]{../figures/rot_dist_900}
-\caption{
+\caption{Color--period diagrams for our sample of likely main sequence stars, divided into bins of distance.
+Our nearest bin (within 350pc) is effectively the distance analyzed in \citet{davenport2017} using Gaia DR1, and clearly shows the rotation period bimodality for the entire sample.
+The brighter magnitude limit of the \Kepler sample results in redder (fainter) stars missing in our further distance bins.
+The rotation period bimodality can be seen in the 350-525 pc bin, but is not found in the bluer stars at further distances.
 }
-%\label{fig:cmd}
+\label{fig:color_period}
 \end{figure*}
 
-\begin{figure}[]
-\centering
-\includegraphics[width=3.5in]{../figures/B_V_rot}
-\caption{
-}
-%\label{fig:cmd}
-\end{figure}
+%\begin{figure}[]
+%\centering
+%\includegraphics[width=3.5in]{../figures/B_V_rot}
+%\caption{converted to B-V using Padova isochrone colors. 3 gyrochrones from \citet{meibom2009} with ages 300 Myr (dark red), 600 Myr (bright red), and 1.2 Gyr (blue) are shown.
+%}
+%%\label{fig:cmd}
+%\end{figure}
+%
 
@@
 \includegraphics[width=3.5in]{../figures/delta_per}
-\caption{
+\caption{Histograms of the log rotation periods after a 600 Myr gyrochrone was subtracted for stars in the same five distance bins shown in Figure \ref{fig:color_period}.
 }
-%\label{fig:cmd}
+\label{fig:per_hist}
 \end{figure}
@@
 }
-%\label{fig:cmd}
+\label{fig:cmd_zoom}
 \end{figure*}

### 2018-05-07 paper/ms.tex
outlined intro

@@
 
-\shorttitle{Rotating Stars}
+\shorttitle{Rotating Stars from \Kepler Observed with Gaia DR2}
 \shortauthors{Davenport et al.}
@@
 \begin{abstract}
-Astrometric data from the recent Gaia Data Release 2 has been matched against the sample of stars from \Kepler with known rotation periods. 
+We have matched the astrometric data from {\em Gaia} Data Release 2 to the sample of stars with measured rotation periods from \Kepler.
+Using 30,305 stars with good distance estimates, we select 16,248 as being likely main sequence single stars within a 0.5 mag region about a 1 Gyr isochrone.
+This removes sub-giants and unresolved binary stars from the sample.
+The rotation period bimodality, originally discovered by \citet{mcquillan2013}, is recovered for stars out to 525pc, but is not detectable at further distances. 
+We also find a significant width in the stellar main sequence of $M_G\sim$0.25 mag, as well as a  increase in the average rotation period correlated with the $M_G$ offset at a given color (mass).
+We interpret this to be the measurable change in luminosity and loss of angular momentum as stars evolve along the main sequence, which may provide a new independent test of stellar evolution and gyrochronlogy models.
+This investigation represents the first step in understanding the star formation history of our solar neighborhood as traced through stellar angular momentum loss. 
 \end{abstract}
@@
 
-The \Kepler mission \citep{borucki2010}
+The \Kepler mission \citep{borucki2010} a new era for enabling detailed study of angular momentum loss in low-mass stars.
+long noted as a means to possibly age-date stars (gyro), open clusters give hope that this general model works for low-mass stars
+many Qs exist about details.
+these include Qs about initial rotation period distribution (CITE), the specific prescription for spin-down (ANGUS, etc), and exploring the efficiency of this angular momentum loss mechanism at old ages (Van Saders).
 
-\citet{mcquillan2013} bimodal period distribution 
-\citep{mcquillan2014}.
+one of the most compelling results from the rotation work in \Kepler is the discovery of a period bimodality. 
+\citet{mcquillan2013} found bimodal feature, in M and later K stars \citep{mcquillan2014}.
+Davenport used Gaia DR1 to remove contamination from sub-giants and found the feature in G dwarfs.
+this feature proposed to be either a new short-lived transition or instability phase of momentum loss, or a signature of star formation history imprinted in the present-day rotation period distribution.
+However, to date this feature has only been observed in the \Kepler rotation period catalog, and most critically only for stars within $\sim$300 pc.
 
-\citet{davenport2017} looked at this with Gaia DR1, extending period gap to G dwarfs, pointing to SFH as cause.
+In this letter we follow the work of \citet{davenport2017} in studying the \Kepler rotation period sample using astrometric data from the {\em Gaia} mission \citep{gaia}.
+By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2, we can use precise distances for essentially every star to select likely main sequence dwarfs to distances well over 1 kpc.
+Importantly this filters out both sub-giants (the main contaminant noted by \cite{davenport2017}), and unresolved binary stars.
+Here we demonstrate the power of such a combined time-domain and astrometric sample for constraining the detailed evolution of main sequence stars themselves, and exploring the star formation history of the Milky Way.
+
+%\citet{davenport2017} looked at this with Gaia DR1, extending period gap to G dwarfs, pointing to SFH as cause.
 sub giant contamination was the thing.
@@
 
- \citep{gaia} mission
-
-In this letter
-
 %%%%%%%%%%%%%%%%%%%%%%
 \section{The \Kepler--Gaia Data}
+
 %Rotation periods in this study come from \citet{mcquillan2014}, who performed an Auto-Correlation Function analysis of \Kepler stars cooler than 6500 K that had at least $\sim$2 years of observation.
@@
 \caption{Histograms of the log rotation periods after a 600 Myr gyrochrone was subtracted for stars in the same five distance bins shown in Figure \ref{fig:color_period}.
+The period bimodality for stars within 350 pc has two peaks similar to those found in \citet{davenport2017}, at -0.15 and +0.18 dex.
+The fast rotating peak (left side) declines sharply at further distances, however.
 }
@@
 \includegraphics[width=6.5in]{../figures/cmd_zoom}
-\caption{
+\caption{Left: Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with MIST isochrones at ages of $10^8$, $10^9$, and $10^10$ yr for comparison (blue, black, and red lines).
+Points are colored by their measured \Kepler rotation periods from \citet{mcquillan2014}.
+The main sequence shows significant scatter, as well as a gradient in rotation period.
+Right: Difference in $M_G$ from the $10^9$ yr MIST isochrone, shown as a function of their rotation periods (point color again indicates rotation period).
+Blue squares show an increase in the median $M_G$ offset in bins of increasing rotation period.
+Error bars shown are the standard deviation of $\Delta M_G$ in each bin.
 }

### 2018-05-09 paper/ms.tex
some words

@@
 many Qs exist about details.
-these include Qs about initial rotation period distribution (CITE), the specific prescription for spin-down (ANGUS, etc), and exploring the efficiency of this angular momentum loss mechanism at old ages (Van Saders).
+these include Qs about initial rotation period distribution \citep[e.g.][]{barnes2010,matt2015}, the specific prescription for spin-down \citep{angus2015}, and exploring the efficiency of this angular momentum loss mechanism at old ages \citep{van-saders2016}.
 
@@
 In this letter we follow the work of \citet{davenport2017} in studying the \Kepler rotation period sample using astrometric data from the {\em Gaia} mission \citep{gaia}.
-By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2, we can use precise distances for essentially every star to select likely main sequence dwarfs to distances well over 1 kpc.
+By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2 \citep{gaia_dr2}, we can use precise distances for essentially every star to select likely main sequence dwarfs to distances well over 1 kpc.
 Importantly this filters out both sub-giants (the main contaminant noted by \cite{davenport2017}), and unresolved binary stars.
 Here we demonstrate the power of such a combined time-domain and astrometric sample for constraining the detailed evolution of main sequence stars themselves, and exploring the star formation history of the Milky Way.
-
-%\citet{davenport2017} looked at this with Gaia DR1, extending period gap to G dwarfs, pointing to SFH as cause.
-sub giant contamination was the thing.
-Now we can finally follow up on this and study the full \Kepler rotation period sample
 
@@
 
-data xmatched by M. Bedell between Kepler DR25 (CITE) and Gaia DR2 (CITE), using 1arcsec radius.
+Our 
+data xmatched by M. Bedell between \Kepler and Gaia DR2 (CITE). using 1arcsec radius, which included 195,830 sources with stellar properties from the \Kepler Data Release 25 and {\em Gaia} DR2.
+
+sub-set was matched to the rotation catalog of \citet{mcquillan2014}
+
+To select stars with good parallaxes, as well as high quality photometry from {\em Gaia}, we selected stars with the following criteria:
+\begin{itemize}
+\item Parallax error $< 0.1$ mas
+\item $\sigma(M_{G}) / M_{G} < 0.01$
+\item $\sigma(G_{BP}) /G_{BP} < 0.01$
+\item $\sigma(G_{RP}) /G_{RP} < 0.01$
+\end{itemize}
+
 distances using the updated prescription from \citet{bailer-jones2018}
+Following the suggested use of \citet{bailer-jones2018}, we use only sources with {\tt modality\_flag} == 1 (i.e.
+not a bimodal distance solution) and {\tt result\_flag} == 1 (i.e.
+well constrained distance).
+
+Our final sample contained 30,305 stars in Gaia DR2 with measured \Kepler rotation periods that passed these selection criteria.
+A color--magnitude diagram of this sample is presented in Figure \ref{fig:cmd}, with points colored by their rotation periods.
+
+\begin{figure}[]
+\centering
+\includegraphics[width=3.6in]{../figures/cmd}
+\caption{
+Color--magnitude diagram for 30,305 \Kepler stars from the \citet{mcquillan2014} sample that are included in Gaia DR2, colored by their measured rotation period.
+For reference we show a $10^9$ year MIST isochrone (black line) used to select likely main sequence, single stars (black line).
+A track of binary  stars is apparent $\sim$0.75 mag above the main sequence.
+As in \citet{davenport2017}, we find significant contamination of the rotation period sample for bluer stars by sub-giants.}
+\label{fig:cmd}
+\end{figure}
 
@@
 
-\begin{figure*}[]
-\centering
-\includegraphics[width=6in]{../figures/cmd}
-\caption{
-Color--magnitude diagram for 30,305 \Kepler stars from the \citet{mcquillan2014} sample that are included in Gaia DR2, colored by their measured rotation period.
-For reference we show a $10^9$ year MIST isochrone (black line) used to select likely main sequence, single stars (black line).
-A track of binary  stars is apparent $\sim$0.7 mag above the main sequence.
-As in \citet{davenport2017}, we find significant contamination of the rotation period sample for bluer stars by sub-giants.}
-\label{fig:cmd}
-\end{figure*}
+As in \citet{davenport2017}, the color--magnitude diagram shown in Figure \ref{fig:cmd} shows many of the bluer stars in the \citet{mcquillan2014} sample are significantly above the main sequence.
+These are likely subgiant stars, which do not follow the main sequence stars spin-down evolution \citep[e.g.][]{donascimento2012, van-saders2013}.
+Since \citet{davenport2017} found subgiants could obscure the rotation period bimodality for G dwarfs, these must be excluded from our analysis, but encourage future studies to explore the wealth of angular momentum evolution data from these most-main sequence objects.
+
+Beyond the subgiant contamination above the main sequence, we can also see a secondary population of stars in a parallel track above the normal main sequence, which are likely unresolved binary stars.
+The pile-up above the main sequence occurs due to unresolved equal-mass field binaries, which was seen in the {\em Gaia} DR1 data as well \citep{anderson2017}.
+Since these systems may have experienced tidal evolution that could significantly impact their rotation evolution \citep[e.g.][]{lurie2017}, we must also remove these from our analysis.
+Though we do not explore the binary population in any detail here, this sample (perhaps with radial velocity follow-up) may provide useful insight into the tidal evolution of binary stars.
+
+we use the Mesa Isochrones and Stellar Tracks \citep[MIST;][]{MIST} to describe the main sequence stars in Figure \ref{fig:cmd} using [Fe/H] =0.25, and an age of $10^9$ years.
+main seq stars chosen in 0.5 mag window
 
@@
 
-This work made use of the \url{gaia-kepler.fun} crossmatch database created by Megan Bedell.
+This work made use of the \url{gaia-kepler.fun} crossmatch database, created by Megan Bedell.
 
@@
 
-%This research made use of the cross-match service, and the SIMBAD database, provided by CDS, Strasbourg, France.
-%
-%This work has made use of data from the European Space Agency (ESA)
-%mission {\it Gaia} (\url{http://www.cosmos.esa.int/gaia}), processed by
-%the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
-%\url{http://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
-%for the DPAC has been provided by national institutions, in particular
-%the institutions participating in the {\it Gaia} Multilateral Agreement.
+This work has made use of data from the European Space Agency (ESA) mission
+{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
+Data Processing and Analysis Consortium (DPAC,
+\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
+has been provided by national institutions, in particular the institutions
+participating in the {\it Gaia} Multilateral Agreement.
 
-\bibliography{/Users/james/Dropbox/references}
+\bibliography{/Users/davenpj3/Dropbox/references}
 

### 2018-05-14 paper/ms.tex
friday writing:

@@
 
-%Rotation periods in this study come from \citet{mcquillan2014}, who performed an Auto-Correlation Function analysis of \Kepler stars cooler than 6500 K that had at least $\sim$2 years of observation.
-The periods recovered from this approach generally agree very well with those found via Lomb-Scargle Periodograms \citep[e.g.][]{reinhold2013,aigrain2015}.
-Sources with multiple distinct periods, such as from binary systems with two spotted stars \citep[e.g.][]{lurie2015} are detected by \citet{mcquillan2014}, but are not included in the following analysis.
+We used the largest homogeneous catalog of rotation periods available from the \Kepler mission.
+The sample from \citet{mcquillan2014} provides rotation periods for more 34,030 stars, measured using the Auto-Correlation Function (ACF).
+While the ACF does not have as fine of resolution in recovering periods as compared with methods such as the Lomb-Scargle Periodogram, it is more robust to detecting the true period as opposed to an alias, and more complete for batch analysis of all stars \citep[e.g.
+see][]{aigrain2015}.
 
-%The Gaia Data Release 1 (DR1) provides astrometric positions for over $10^9$ sources from the first year of observation with Gaia.
-The Tycho-Gaia Astrometric Solution (TGAS) measures improved proper motions and parallaxes for 2 million nearby, bright sources by extending the astrometric solutions from Tycho and Hipparcos.
-While the TGAS data are not a complete astrometric survey, and have possible systematics in the reported parallaxes \citep{gaia_dr1,stassun2016}, they represent a significant improvement in the astrometry and kinematics available for stars in the \Kepler field.
-
-Our 
-data xmatched by M. Bedell between \Kepler and Gaia DR2 (CITE). using 1arcsec radius, which included 195,830 sources with stellar properties from the \Kepler Data Release 25 and {\em Gaia} DR2.
-
-sub-set was matched to the rotation catalog of \citet{mcquillan2014}
+The \Kepler data was matched to the {\em Gaia} DR2 source catalog using a 1 arcsecond radius.
+We used the \Kepler--Gaia cross-match made publicly available by M.
+Bedell, which included entries for 195,830 sources.
+\Kepler-based stellar parameters included in this cross-match come the Data Release 25 \Kepler catalog.
+Joining this cross-matched table to the \citet{mcquillan2014} catalog, we found 33,538 sources.
 
@@
 
-distances using the updated prescription from \citet{bailer-jones2018}
-Following the suggested use of \citet{bailer-jones2018}, we use only sources with {\tt modality\_flag} == 1 (i.e.
+Rather than simply use the inverse {\em Gaia} parallax values to measure the distance to sources, we use the improved distance prescription from \citet{bailer-jones2018}, who provided independent distances estimates for 1.33 billion {\em Gaia} sources using a weak prior on the distribution of stars in our galaxy.
+We follow their suggested use of the distance catalog, we only include sources with {\tt modality\_flag} == 1 (i.e.
 not a bimodal distance solution) and {\tt result\_flag} == 1 (i.e.

### 2018-05-17 paper/ms.tex
some very interesting words...

@@
 
-As in \citet{davenport2017}, the color--magnitude diagram shown in Figure \ref{fig:cmd} shows many of the bluer stars in the \citet{mcquillan2014} sample are significantly above the main sequence.
+As in \citet{davenport2017}, the color--magnitude diagram in Figure \ref{fig:cmd} shows many of the bluer stars in the \citet{mcquillan2014} sample are located significantly above the main sequence.
 These are likely subgiant stars, which do not follow the main sequence stars spin-down evolution \citep[e.g.][]{donascimento2012, van-saders2013}.
-Since \citet{davenport2017} found subgiants could obscure the rotation period bimodality for G dwarfs, these must be excluded from our analysis, but encourage future studies to explore the wealth of angular momentum evolution data from these most-main sequence objects.
+Since \citet{davenport2017} found subgiants could obscure the rotation period bimodality for G dwarfs, these must be excluded from our analysis, but we encourage future studies to explore the wealth of angular momentum evolution data from these most-main sequence objects.
 
-Beyond the subgiant contamination above the main sequence, we can also see a secondary population of stars in a parallel track above the normal main sequence, which are likely unresolved binary stars.
-The pile-up above the main sequence occurs due to unresolved equal-mass field binaries, which was seen in the {\em Gaia} DR1 data as well \citep{anderson2017}.
+Beyond the subgiant contamination, we also see a secondary population of stars in a parallel track $\sim$0.75 mag above the normal main sequence, which are attributed to unresolved binary star systems.
+This pile-up above the main sequence occurs due to unresolved equal-mass (or nearly equal-mass) field binaries, and was seen in the {\em Gaia} DR1 data as well \citep{anderson2017}.
 Since these systems may have experienced tidal evolution that could significantly impact their rotation evolution \citep[e.g.][]{lurie2017}, we must also remove these from our analysis.
-Though we do not explore the binary population in any detail here, this sample (perhaps with radial velocity follow-up) may provide useful insight into the tidal evolution of binary stars.
+Though we do not explore the binary population in any detail here, this sample (perhaps with radial velocity follow-up) may provide useful insight into the tidal evolution of binary stars, and are good targets for characterizing binary star system properties.
+We also note a small number of systems above the even the equal-mass binary main sequence track, which could be due to unresolved triple star systems.
 
-we use the Mesa Isochrones and Stellar Tracks \citep[MIST;][]{MIST} to describe the main sequence stars in Figure \ref{fig:cmd} using [Fe/H] =0.25, and an age of $10^9$ years.
-main seq stars chosen in 0.5 mag window
+We use an isochrone from the Mesa Isochrones and Stellar Tracks suite \citep[MIST;][]{MIST} to choose likely main sequence stars in Figure \ref{fig:cmd}.
+Our favored model to represent the main sequence in this study had [Fe/H] = +0.25 and an age of $10^9$ years, and was chosen by-hand.
+Single, main sequence stars were selected in a region spanning 0.1 mag fainter and 0.4 mag brighter than the MIST isochrone, resulting in a final sample  of 16,248 stars for analysis of their rotation period distributions.
 
-%Though \citet{mcquillan2014} attempted to only measure periods for dwarf stars, the sample of \Kepler--Gaia matched stars contains both main sequence dwarfs and evolved stars (giants and subgiants).
-Previous studies have shown  significant contamination by giants or subgiants can affect the implied variability properties of dwarf stars \citep{ciardi2011,mann2012}.
-Therefore to properly understand the nature of the period distribution and its implications for age-dating field stars, a robust sample of main sequence stars must be selected.
-%
-%Faint stars were removed by requiring sources have $G$-band flux errors $<1$\%.
-To ensure accurate distances, and therefore luminosities, parallaxes were required to have errors $<0.4$ mas.
-These cuts left a total of 894 stars from the \Kepler--TGAS matched sample (68\%).
-The Hertzsprung--Russell (HR) diagram for these stars is shown in Figure \ref{fig:HR}, with each point colored by its \Kepler-measured rotation period.
-Example isochrones from the \citet{bressan2012} grid are shown for two ages.
-A systematic offset of $\sim$0.5 magnitudes is found between the measured absolute $G$-band and the isochrone's main sequence.
-This offset is likely due to calibration differences between the nominal and actual $G$-band (A.
-Brown, private communication).
-%
-%
-%The HR-period diagram in Figure \ref{fig:HR} shows stars with a range of evolutionary states, and could help test post-main sequence angular momentum evolution models \citep[e.g.][]{donascimento2012}.
-Outliers in this diagram are either due to erroneous cross-matching in the \Kepler and Gaia catalogs, or represent interesting systems such as rare binary star configurations or stars that have undergone mergers or ingested giant planets \citep{massarotti2008,tayar2015}.
-For example, examining the 2 Micron All Sky Survey \citep[2MASS][]{2mass} image on SIMBAD for
-%the rapidly rotating star at $T_{eff}=4500$ K, $M_G=4.5$ mag, $P_{rot}$=0.652 d, in Figure \ref{fig:HR} (KIC 07957709), this source appears highly contaminated with three point sources clustered within $\sim$10 arcseconds, likely leading to an erroneous position in the HR diagram, and possibly incorrect variability measurements with \Kepler.
-%Investigating all such outliers in Figure \ref{fig:HR} is beyond the scope of this work, but these targets are worth further study as they may reveal new physics.
-%
-%
-%
-%Main sequence stars were selected using a simple cut around 300 Myr isochrone.
-Given the systematic offset of $\sim$0.5 mag between the Myr isochrone and the observed $M_G$ values, a fairly wide band of stars ($0\le \Delta M_G \le 1$) was selected as being ``close to the main sequence''.
-The final sample included 440 stars.
-This simplistic cut is not a robust dwarf--giant, nor single--binary star separator, but serves to select a sample of mostly main sequence stars for the illustrative purpose of this work.
-More precise selection will require an improved isochrone track, as well as updated parallaxes from the full Gaia DR2.
-%
+%%%%%%%%%%%%%%%%%%%%%%
+\section{Tracing the Period Bimodality}
 
-%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Extending the Spin-Down Gap}
-%A bimodal period distribution was first discovered by \citet{mcquillan2013} for \Kepler M dwarfs, who found a dearth of objects with periods around $\sim$25 days.
-Follow-up work by \citet{mcquillan2014} found this bimodality extended to K dwarfs, up to $T_{eff}\sim5500$.
-Figure \ref{fig:gyro} shows the rotation period distribution for the final 440 star sample of likely main sequence stars.
-The bimodality appears to extend smoothly through to the hottest stars in this sample.
-%
-%While these periods for field stars were robustly measured by \citet{mcquillan2014} and others, the bimodality did not appear in previous \Kepler work due to the high contamination rate by subgiants for these bluer, hotter stars.
-414 rotating stars had acceptable photometric and parallax uncertainties in TGAS, but were culled from this sample for having $M_G$ luminosities higher than the main sequence cut in \S3 above.
-Note that distributions of $\log g$ values from the Kepler Input Catalog \citep{brown2011a} for both the main sequence and subgiant stars were not statistically different.
+all the color--period diagrams vs distance:
 
@@
 
-%\begin{figure}[]
-%\centering
-%\includegraphics[width=3.5in]{../figures/B_V_rot}
-%\caption{converted to B-V using Padova isochrone colors. 3 gyrochrones from \citet{meibom2009} with ages 300 Myr (dark red), 600 Myr (bright red), and 1.2 Gyr (blue) are shown.
-%}
-%%\label{fig:cmd}
-%\end{figure}
-%
+re-define gyrochrone in terms of 
+subtract 600myr gyrochrone off again, as in D17
 
@@
 \centering
-\includegraphics[width=3.5in]{../figures/delta_per}
+\includegraphics[width=3.5in]{../figures/delta_per_2gauss}
 \caption{Histograms of the log rotation periods after a 600 Myr gyrochrone was subtracted for stars in the same five distance bins shown in Figure \ref{fig:color_period}.
@@
 
-\begin{figure*}[]
+these distance bins reach approximate distances above the galactic midplane of $\sim$60 pc in the nearest bin, to nearly 800 pc in the furthest bin, therefore spanning a large range in apparent age and possible star formation history
+%  61.70587735  102.55881602  120.06721831  190.10082747  797.0587735  pc
+
+\clearpage
+%%%%%%%%%%%%%%%%%%%%%%
+\section{Recalibrating Stellar Evolution Models}
+
+\begin{figure*}
 \centering
-\includegraphics[width=6.5in]{../figures/cmd_zoom}
-\caption{Left: Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with MIST isochrones at ages of $10^8$, $10^9$, and $10^10$ yr for comparison (blue, black, and red lines).
-Points are colored by their measured \Kepler rotation periods from \citet{mcquillan2014}.
-The main sequence shows significant scatter, as well as a gradient in rotation period.
-Right: Difference in $M_G$ from the $10^9$ yr MIST isochrone, shown as a function of their rotation periods (point color again indicates rotation period).
+\includegraphics[width=6in]{../figures/cmd_zoom}
+\caption{
+Left: Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with stars are colored by their measured \Kepler rotation periods from \citet{mcquillan2014}.
+MIST isochrones at ages of $10^8$, $10^9$, and $10^10$ yr are shown for comparison (blue, black, and red lines).
+The predicted evolution of a 0.7 M$_\odot$ star is highlighted (black squares).
+The main sequence shows significant scatter, as well as a gradient in rotation period with slower rotators preferentially above the main sequence.
+Right: Difference in $M_G$ from the $10^9$ yr MIST isochrone, shown as a function of their rotation periods (point color again indicates rotation period, as in Left panel).
 Blue squares show an increase in the median $M_G$ offset in bins of increasing rotation period.
@@
 
-%\begin{figure}[]
-%\centering
-%\includegraphics[width=3.5in]{../figures/iso_age}
-%\caption{
-%}
-%%\label{fig:cmd}
-%\end{figure}
+A subtle feature that we noticed in Figure \ref{fig:cmd} is the apparent color gradient (i.e.
+rotation period gradient) between the single and binary star main sequence populations.
+In Figure \ref{fig:cmd} this appears as a yellow stripe (i.e.
+rotation periods of 30-40 days) between these blue-green sequences for systems with colors of $G_{BP} - G_{RP} \approx 1.5$.
+To exaggerate this feature, we have reproduced a portion of our color--magnitude diagram focused on the main sequence near this stellar color in the left panel of Figure \ref{fig:cmd_zoom}.
+A clear color gradient is present, with red points (slower rotators) appearing preferentially above the main sequence.
+The right panel of Figure \ref{fig:cmd_zoom} concretely demonstrates a correlation between the measured rotation period and the vertical offset (i.e.
+absolute magnitude) from the $10^9$ year MIST isochrone.
+Slower rotating stars are apparently brighter at a given stellar color.
+
+We interpret this observation as being due to the increase in luminosity of stars across their main sequence lives, coupled with the angular momentum loss that underpins gyrochronology.
+Indeed, the sequence of MIST isochrones shown in Figure \ref{fig:cmd_zoom} with ages of $10^8$, $10^9$, and $10^10$ years clearly predicts an increase in brightness of the main sequence over time.
+
+indeed, 0.7solarmass star traced over its life in both panels doesnt quite make sense. 
+inverting CMD into precise ages w/ Gaia beyond scope of this work, but this figure suggests the MS is too wide given a reasonable age range.
+consider: younger stars, presumably more metal rich, should be redder, moving in the wrong direction to the affect we see here.
+more colors and careful work needed...
+vertical action is OK for a star, but maybe some color impact we not considering....
+hmm.
+
+however, looking at 0.7M period vs lum evolution, we see that the period seems to also drop off too rapidly compared to a model.
+combining the MM09 spin-down formulae for a 0.7M star, with the MIST isochrone evolution, we see  the period range should go out to 60--80 days to get the right amplitude offset.
+That implies that at a given age (or vertical action), the stars are rotating {\it faster} than expected.
+this is similar to the broken spin-down model suggested by \citep{van-saders2016}, and is potentially the first confirmation of this observation.
 
@@
 
-\bibliography{/Users/davenpj3/Dropbox/references}
+\bibliography{/Users/james/Dropbox/references}
 

### 2018-05-19 paper/ms.tex
a couple words

@@
 MIST isochrones at ages of $10^8$, $10^9$, and $10^10$ yr are shown for comparison (blue, black, and red lines).
-The predicted evolution of a 0.7 M$_\odot$ star is highlighted (black squares).
-The main sequence shows significant scatter, as well as a gradient in rotation period with slower rotators preferentially above the main sequence.
-Right: Difference in $M_G$ from the $10^9$ yr MIST isochrone, shown as a function of their rotation periods (point color again indicates rotation period, as in Left panel).
+The predicted evolution of a 0.7 M$_\odot$ star is highlighted (black triangles).
+%The main sequence shows significant scatter, as well as a gradient in rotation period with slower rotators preferentially above the main sequence.
+Right: Difference in $M_G$ from the $10^9$ yr MIST isochrone as a function of rotation period (point color again indicates rotation period, as in Left panel).
 Blue squares show an increase in the median $M_G$ offset in bins of increasing rotation period.
 Error bars shown are the standard deviation of $\Delta M_G$ in each bin.
+The 0.7 M$_\odot$ star brightness evolution is shown as a function of rotation evolution from \citet{meibom2009} (black triangles).
 }

### 2018-05-21 paper/ms.tex
a couple words

@@
 consider: younger stars, presumably more metal rich, should be redder, moving in the wrong direction to the affect we see here.
-more colors and careful work needed...
-vertical action is OK for a star, but maybe some color impact we not considering....
+more stellar masses and careful work needed...
+vertical spread in this parameter space is OK for a star, but maybe some color impact we not considering....
 hmm.
@@
 combining the MM09 spin-down formulae for a 0.7M star, with the MIST isochrone evolution, we see  the period range should go out to 60--80 days to get the right amplitude offset.
-That implies that at a given age (or vertical action), the stars are rotating {\it faster} than expected.
+That implies that at a given age (or vertical offset), the stars are rotating {\it faster} than expected.
 this is similar to the broken spin-down model suggested by \citep{van-saders2016}, and is potentially the first confirmation of this observation.

### 2018-05-22 paper/ms.tex
lots of words today

@@
 \section{Tracing the Period Bimodality}
+Using this sample of likely single, main sequence stars from \Kepler and {\em Gaia}, we are able to explore the distribution of rotation periods for stars as a function of their distance.
+The rotation period bimodality in \Kepler stars previously only detected for stars within $\sim$300 pc of the Sun due to the limits of available parallax data.
+Now with {\em Gaia} DR2, our sample of main sequence star with measured rotation periods in \Kepler extends out to over 2 kpc.
 
-all the color--period diagrams vs distance:
+In Figure \ref{fig:color_period} we present the period--color diagram for our sample of stars, split into six bins of projected distance.
+The first panel (0--350 pc) effectively reproduces the results of \citet{davenport2017} for bluer stars and \citet{mcquillan2014} for the redder stars.
+A gap in the observed rotation periods as a function of color is seen, at a period of approximately 5 days for $G_{BP}-G_{RP}\approx1$, 20 days for $G_{BP}-G_{RP}\approx2$, and increasing towards 30 days for the reddest stars in our sample.
+This gap corresponds with a line of approximately constant age, consistent with a gyrochrone with age $\sim$600 Myr \citep{davenport2017}.
 
@@
 \includegraphics[width=3.5in]{../figures/rot_dist_600}
+\includegraphics[width=3.5in]{../figures/rot_dist_700}
 \includegraphics[width=3.5in]{../figures/rot_dist_900}
-\caption{Color--period diagrams for our sample of likely main sequence stars, divided into bins of distance.
+\caption{Color--period diagrams for our sample of likely main sequence stars, divided into six bins of distance.
 Our nearest bin (within 350pc) is effectively the distance analyzed in \citet{davenport2017} using Gaia DR1, and clearly shows the rotation period bimodality for the entire sample.
@@
 
-re-define gyrochrone in terms of 
-subtract 600myr gyrochrone off again, as in D17
+The bimodality is still apparent in the second distance bin (350--525 pc), clearly visible in the redder stars, but seems to fade in the final three bins.
+Other structures in the period distributions are visible, however.
+For example, a thin sequence of stars with rotation periods near 10 days is faintly visible in the most distant bin (900--2500 pc) for stars with colors of $0.8<G_{BP}-G_{RP}<1.2$.
+This feature is due to the 1 Gyr open cluster NGC 6811 in the \Kepler field \citep{meibom2011}, whose distance is $\sim$1100 pc \citep{sandquist2016}.
+
+To better illustrate the evolution of rotation periods for all the stars between distance bins, in Figure \ref{fig:per_hist} we follow \citet{davenport2017} and subtract the rotation period of a 600 Myr gyrochrone.
+As no published gyrochronology model yet exists that has been tuned to the {\em Gaia} photometric colors, we adopt the same gyrochronology model of Eqn.
+2 from \citet{meibom2009} used by \citet{davenport2017} to approximately trace the rotation period gap at 600 Myr as a function of $B-V$ color.
+We convert stars from the observed {\em Gaia} $G_{BP}-G_{RP}$ color to $B-V$ using the same 1 Gyr MIST isochrone model used to define the main sequence in Figure \ref{fig:cmd} above.
+
+Our sample naturally becomes biased towards the bluer (brighter) stars as we reach larger distances.
+Since bluer stars have more rapid rotation and more dramatic period evolution, gyrochronology models are ``bent'' strongly for these stars, and thus the period bimodality (or other structures) become difficult to distinguish.
+\citet{davenport2017} used an ad-hoc modification of the gyrochrone model for the hottest stars to illustrate the existence of the period bimodality.
+However, in Figure \ref{fig:per_hist} we limit our analysis to stars with $0.8<B-V<1.5$ (approximately $0.9<G_{BP}-G_{RP}<2.2$, or $0.9>M_\odot> 0.5$), where the gyrochrone is most flat as a function of color.
+This color range was chosen to ensure ample stars were available in each distance bin, but without having to use the ad-hoc gyrochrone model correction from \citet{davenport2017}.
+
+Figure \ref{fig:per_hist} shows clearly that for the rotation period bimodality dominates for the nearest stars
 
@@
 \includegraphics[width=3.5in]{../figures/delta_per_2gauss}
-\caption{Histograms of the log rotation periods after a 600 Myr gyrochrone was subtracted for stars in the same five distance bins shown in Figure \ref{fig:color_period}.
-The period bimodality for stars within 350 pc has two peaks similar to those found in \citet{davenport2017}, at -0.15 and +0.18 dex.
-The fast rotating peak (left side) declines sharply at further distances, however.
+\caption{Histograms of the log rotation periods after a 600 Myr gyrochrone was subtracted for stars in the same six distance bins shown in Figure \ref{fig:color_period} (faint lines).
+Two-gaussian models were fit to each histogram (bold lines).
+The period bimodality for stars within 350 pc has two nearly equal peaks similar to those found in \citet{davenport2017}, at -0.15 and +0.18 dex.
+The fast rotating peak (left side) declines sharply at further distances.
 }
@@
 
+\Kepler field has a central Galactic latitude of $b\sim13.5^\circ$, pointing slightly above the Galactic midplane. 
 these distance bins reach approximate distances above the galactic midplane of $\sim$60 pc in the nearest bin, to nearly 800 pc in the furthest bin, therefore spanning a large range in apparent age and possible star formation history
 %  61.70587735  102.55881602  120.06721831  190.10082747  797.0587735  pc
+Both galaxy formation simulations \citep{ma2017} and observations of stars in the nearby Milky Way \citep{xiang2017} indicate that height above the midplane correlates strongly with the median ages for stars.
+However, since the \Kepler field is not very upwardly inclined, we not reaching significant {\it heights} above the plane.
+therefore we think the localization of the period bimodality is related to an age feature that is localized within the disk near the Sun, not representative of the total star formation history of our galaxy.
 
-\clearpage
+\citet{lewis2015} for example show lots of variation in SFH in 100pc bins for M31
+
+%\clearpage
 %%%%%%%%%%%%%%%%%%%%%%

### 2018-05-22 paper/ms.tex
good progress today

@@
 
-In this letter we follow the work of \citet{davenport2017} in studying the \Kepler rotation period sample using astrometric data from the {\em Gaia} mission \citep{gaia}.
+In this paper we follow the work of \citet{davenport2017} in studying the \Kepler rotation period sample using astrometric data from the {\em Gaia} mission \citep{gaia}.
 By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2 \citep{gaia_dr2}, we can use precise distances for essentially every star to select likely main sequence dwarfs to distances well over 1 kpc.
@@
 \section{Tracing the Period Bimodality}
-Using this sample of likely single, main sequence stars from \Kepler and {\em Gaia}, we are able to explore the distribution of rotation periods for stars as a function of their distance.
-The rotation period bimodality in \Kepler stars previously only detected for stars within $\sim$300 pc of the Sun due to the limits of available parallax data.
-Now with {\em Gaia} DR2, our sample of main sequence star with measured rotation periods in \Kepler extends out to over 2 kpc.
-
-In Figure \ref{fig:color_period} we present the period--color diagram for our sample of stars, split into six bins of projected distance.
-The first panel (0--350 pc) effectively reproduces the results of \citet{davenport2017} for bluer stars and \citet{mcquillan2014} for the redder stars.
-A gap in the observed rotation periods as a function of color is seen, at a period of approximately 5 days for $G_{BP}-G_{RP}\approx1$, 20 days for $G_{BP}-G_{RP}\approx2$, and increasing towards 30 days for the reddest stars in our sample.
-This gap corresponds with a line of approximately constant age, consistent with a gyrochrone with age $\sim$600 Myr \citep{davenport2017}.
 
@@
 
+Using this sample of likely single, main sequence stars from \Kepler and {\em Gaia}, we are able to explore the distribution of rotation periods for stars as a function of their distance.
+The rotation period bimodality in \Kepler stars previously only detected for stars within $\sim$300 pc of the Sun due to the limits of available parallax data.
+Now with {\em Gaia} DR2, our sample of main sequence star with measured rotation periods in \Kepler extends out to over 2 kpc.
+
+In Figure \ref{fig:color_period} we present the period--color diagram for our sample of stars, split into six bins of projected distance.
+The first panel (0--350 pc) effectively reproduces the results of \citet{davenport2017} for bluer stars and \citet{mcquillan2014} for the redder stars.
+A gap in the observed rotation periods as a function of color is seen, at a period of approximately 5 days for $G_{BP}-G_{RP}\approx1$, 20 days for $G_{BP}-G_{RP}\approx2$, and increasing towards 30 days for the reddest stars in our sample.
+This gap corresponds with a line of approximately constant age, consistent with a gyrochrone with age $\sim$600 Myr \citep{davenport2017}.
+
 The bimodality is still apparent in the second distance bin (350--525 pc), clearly visible in the redder stars, but seems to fade in the final three bins.
@@
 
-To better illustrate the evolution of rotation periods for all the stars between distance bins, in Figure \ref{fig:per_hist} we follow \citet{davenport2017} and subtract the rotation period of a 600 Myr gyrochrone.
+\begin{figure*}[!t]
+\centering
+\includegraphics[width=3.5in]{../figures/delta_per_2gauss}
+\includegraphics[width=3.5in]{../figures/delta_per_2gauss_Z}
+\caption{Left: Distribution of the log rotation periods after a 600 Myr gyrochrone was subtracted, using the same distance bins shown in Figure \ref{fig:color_period} (faint lines).
+Two-gaussian models were fit to each histogram (bold lines).
+The rotation period bimodality for stars within 350 pc has two nearly equal peaks similar to those found in \citet{davenport2017}, at -0.15 and +0.18 dex.
+The fast rotating peak (left side) declines sharply at further distances.
+Right: Same as Left, but in bins of Galactic height above the plane (Z). The drop-off of fast rotating, young stars is even more pronounced as a function of height.
+}
+\label{fig:per_hist}
+\end{figure*}
+
+To better illustrate the evolution of rotation periods for all the stars between the distance bins, in Figure \ref{fig:per_hist} we follow \citet{davenport2017} and subtract the rotation period of a 600 Myr gyrochrone.
 As no published gyrochronology model yet exists that has been tuned to the {\em Gaia} photometric colors, we adopt the same gyrochronology model of Eqn.
@@
 
-Figure \ref{fig:per_hist} shows clearly that for the rotation period bimodality dominates for the nearest stars
+Figure \ref{fig:per_hist} (left) shows clearly that for the short period component of the rotation period bimodality dominates only for the nearest stars, and steadily decreases with increasing distance.
+%but can also cut vs Z instead of d. See bimodality drops neatly with height in Figure \ref{fig:per_hist} also.
+The \Kepler field is centered on a Galactic latitude of $b\sim13.5^\circ$, and so we can also study the period bimodality as a function of height above the Galactic mid-plane.
+Both galaxy formation simulations \citep{ma2017} and observations of stars in the nearby Milky Way \citep{xiang2017} indicate that height above the mid-plane correlates strongly with the median ages for stars, out to distances of several kpc.
+The Milky Way thin disk has a scaleheight of $\sim$300 pc \citep{gilmore1983}.
+However, since the \Kepler field is oriented towards low latitudes we do not reach significant {\it heights} above the disk.
+The projected height above the mid-plane for the distance bins shown in Figures \ref{fig:color_period} and \ref{fig:per_hist} ranges from $Z\sim100$ pc at a distance of $d=350$ pc to $Z\sim230$ pc at $d=900$ pc.
+As a result, we are only sensitive to changes in the youngest stars within this span of $Z$.
 
-\begin{figure}[]
+In Figure \ref{fig:per_hist} (right) we find that the drop-off of the short-period (rapid rotating) component of the period bimodality is even more pronounced as a function of $Z$, decreasing rapidly after only 90 pc.
+With increasing height we also see that the shift to the longer period component is more smooth.
+However, from these two projections in Figure \ref{fig:per_hist} alone (distance and height) we cannot definitively determine the spatial structure of the rotation period bimodality.
+
+As a first attempt at understanding the spatial extent of this age-related feature, in Figure \ref{fig:dZ} we break the lowest height stars ($Z < 100$ pc) in to two roughly even samples, split as a function of their projected distance.
+Note we have also repeated this exercise for stars in higher ranges of $Z$, and find the decline of the rapid rotators is again uniform between subsamples of varying projected distance.
+The rotation period bimodality is clearly seen in both distance bins of Figure \ref{fig:dZ}, indicating that the feature is likely not a localized star formation history artifact centered around the Sun.
+Instead, we believe this feature is characteristic of the age--$Z$ dynamical correlation observed within the solar neighborhood of our Galaxy.
+
+\begin{figure}[!ht]
 \centering
-\includegraphics[width=3.5in]{../figures/delta_per_2gauss}
-\caption{Histograms of the log rotation periods after a 600 Myr gyrochrone was subtracted for stars in the same six distance bins shown in Figure \ref{fig:color_period} (faint lines).
-Two-gaussian models were fit to each histogram (bold lines).
-The period bimodality for stars within 350 pc has two nearly equal peaks similar to those found in \citet{davenport2017}, at -0.15 and +0.18 dex.
-The fast rotating peak (left side) declines sharply at further distances.
+\includegraphics[width=3.5in]{../figures/dist_Z}
+\includegraphics[width=3.5in]{../figures/delta_per_subZ}
+\caption{
+Top: projected distance and Galactic height of the main sequence stars in our sample.
+Two approximately equal subsets of stars near the Galactic mid-plane have been highlighted, those with distances from the sun of $d<300$ pc (red), and with $300<d<525$ pc.
+Bottom: As in Figure \ref{fig:per_hist}, the distribution of rotation periods after a 600 Myr gyrochrone was subtracted (faint lines), and with two-Gaussian fits (bold lines) for the same subsets of stars as above.
+The period bidmodality is seen in both distance bins, suggesting it is not localized around the Sun.
 }
-\label{fig:per_hist}
+\label{fig:dZ}
 \end{figure}
 
-\Kepler field has a central Galactic latitude of $b\sim13.5^\circ$, pointing slightly above the Galactic midplane. 
-these distance bins reach approximate distances above the galactic midplane of $\sim$60 pc in the nearest bin, to nearly 800 pc in the furthest bin, therefore spanning a large range in apparent age and possible star formation history
-%  61.70587735  102.55881602  120.06721831  190.10082747  797.0587735  pc
-Both galaxy formation simulations \citep{ma2017} and observations of stars in the nearby Milky Way \citep{xiang2017} indicate that height above the midplane correlates strongly with the median ages for stars.
-However, since the \Kepler field is not very upwardly inclined, we not reaching significant {\it heights} above the plane.
-therefore we think the localization of the period bimodality is related to an age feature that is localized within the disk near the Sun, not representative of the total star formation history of our galaxy.
-
-\citet{lewis2015} for example show lots of variation in SFH in 100pc bins for M31
-
+\vspace{1in}
 %\clearpage
@@
 Left: Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with stars are colored by their measured \Kepler rotation periods from \citet{mcquillan2014}.
-MIST isochrones at ages of $10^8$, $10^9$, and $10^10$ yr are shown for comparison (blue, black, and red lines).
+MIST isochrones at ages of $10^8$, $10^9$, and $10^{10}$ yr are shown for comparison (blue, black, and red lines).
 The predicted evolution of a 0.7 M$_\odot$ star is highlighted (black triangles).
@@
 We interpret this observation as being due to the increase in luminosity of stars across their main sequence lives, coupled with the angular momentum loss that underpins gyrochronology.
-Indeed, the sequence of MIST isochrones shown in Figure \ref{fig:cmd_zoom} with ages of $10^8$, $10^9$, and $10^10$ years clearly predicts an increase in brightness of the main sequence over time.
+Indeed, the sequence of MIST isochrones shown in Figure \ref{fig:cmd_zoom} with ages of $10^8$, $10^9$, and $10^{10}$ years clearly predicts an increase in brightness of the main sequence over time.
 
@@
 combining the MM09 spin-down formulae for a 0.7M star, with the MIST isochrone evolution, we see  the period range should go out to 60--80 days to get the right amplitude offset.
-That implies that at a given age (or vertical offset), the stars are rotating {\it faster} than expected.
-this is similar to the broken spin-down model suggested by \citep{van-saders2016}, and is potentially the first confirmation of this observation.
+That implies that at a given age (or vertical offset), the stars are rotating {\it slightly faster} than expected.
+This is qualitatively similar to the broken spin-down at a critical Rossby number model suggested by \citet{van-saders2016}, though we cannot definitively confirm such intriguing rotation evolution from this initial investigation.
+However, more detailed studies of \Kepler and K2 rotating stars matched to {\em Gaia} may be an ideal ensemble dataset to test the \citet{van-saders2016} model against.
 
@@
 \section{Discussion}
-%
-%Using a combination of data from \Kepler and Gaia DR1, I have explored the rotation period distribution for 440 nearby main sequence stars.
-A bimodal rotation period distribution has been found in stars with temperatures ranging from 5000 K to 6500 K.
-This feature matches that found in cooler stars from \Kepler, but was only revealed thanks to the enhanced ability to distinguish dwarfs from subgiants using Gaia data.
-%A tenuous difference in the TGAS total proper motion for stars in the fast and slow rotating groups is found, which is in agreement with the findings for cool stars by \citet{mcquillan2013}.
-%
-%While a definitive explanation for this period bimodality has not been reached, the findings to date seem to favor stellar ages as the cause.
-In this scenario the star formation history for nearby stars would be dominated by two epochs of star formation, one short event centered at a few hundred Myr, and one long event centered at a few Gyr (slightly younger than the Sun).
-It is also worth noting that the space volume probed by the TGAS sample investigated here is very similar to that covered by the temperature-selected cool star sample in \citet{mcquillan2013}.
-The median parallax distance for stars in this work is 285 pc, while the median isochrone distance for the K and M dwarfs is $\sim$216 pc.
-This points to the period distribution being a localized age artifact.
-%Determining how localized this age distribution is, and if it can be confirmed for stars across the HR diagram including giants, is a key goal for future Gaia data releases.
-%
+
+% from paper1
 %
@@
 Such a feature has also been discussed for rotating stars by \citet{kado-fong2016}.
-Given that the mass range for the bimodality explored here and in \citet{mcquillan2014} covers stars with solar-type dynamos (those having a tachocline, late F through early M) such a model cannot be fully ruled out at this time.
-Though there have been many rotation studies for cool stars \citep[e.g.][]{irwin2011,newton2016,stelzer2016} too few rotation periods have been measured for stars across the ``fully convective boundary'' ($T_{eff}<3000$ K, spectral type $\sim$M4) to tell if the bimodal period feature continues to cooler temperatures, which would support the age distribution model.
-%If the bimodality is due to stars crossing a phase of rapid angular momentum evolution, we would expect to see it in stellar clusters at or near the critical age.
-The lack of this feature in the clusters observed to date could be due to no cluster being close enough to the critical age, which the gyrochrone in Figure \ref{fig:gyro} shows is near 600 Myr.
-Further studies of rotation periods for stars in intermediate age open clusters (e.g.
-the Hyades) may help solve this mystery \citep[e.g.][]{douglas2014}.
-%
-%Finally, this exploratory work has highlighted the utility of using astrometric data from Gaia combined with detailed light curve statistics from \Kepler to reveal hidden substructure in the properties of field stars.
-Looking forward to the astrometric precision of future Gaia data releases, this combination will be effective at separating dwarf stars from subgiants for nearly the entire \Kepler and K2 databases, and enable accurate age maps for field stars.
+Given that the mass range for the bimodality explored
+
+we find period bimodality is localized to low galactic heights, not only close distances to the Sun.
+using rotation to trace the ages of nearby stars. would be really neat to compare the spatial structure of star formation in detail, as 
+\citet{lewis2015} for example show lots of variation in SFH in 100pc bins for M31
+%\citet{williams2017} also for M31, older SFH tracers
+and find out what the actual extent/coherance of the bimodality is.
+
+Adding to the sample of rotation periods from Kepler using K2 (Foreman-Mackey in prep?), not only for doubling the sample size to investigate trends with more precision, but also the wider range of galactic fields of view mean wider range of ages/SFH's we can sample.
+
+We note a shortcoming of our approach in normalizing the rotation period distributions to a 600 Myr gyrochrone in Figure \ref{fig:per_hist} is that it gives the false appearance that this age is a universal transition point.
+A more robust approach is needed that uses improved gyrochronology models to invert the observed color--period diagrams into age distributions.
+
+finally, this work has found strong motivation for a new generation of gyrochronology isochrones to adequately model the {\em Gaia}--\Kepler/K2 (and soon TESS) rotation data for field stars.
+As we show in Figure \ref{fig:cmd_zoom}, these models should include both the angular momentum evolution (period), as well as the temperature and luminosity (color, magnitude) as a function of time.
+With such improvements, we may be able to
 

### 2018-06-25 paper/ms.tex
tweaking some figures
section1 edited

@@
 \bibliographystyle{aasjournal}
+
 \usepackage{graphicx}
@@
 \shorttitle{Rotating Stars from \Kepler Observed with Gaia DR2}
-\shortauthors{Davenport et al.}
+\shortauthors{Davenport \& Covey}
 
@@
 We have matched the astrometric data from {\em Gaia} Data Release 2 to the sample of stars with measured rotation periods from \Kepler.
-Using 30,305 stars with good distance estimates, we select 16,248 as being likely main sequence single stars within a 0.5 mag region about a 1 Gyr isochrone.
-This removes sub-giants and unresolved binary stars from the sample.
+Using 30,305 stars with good distance estimates, we select 16,248 as being likely main sequence single stars centered within a 0.5 mag region about a 1 Gyr isochrone.
+This removes many sub-giants and unresolved binary stars from the sample.
 The rotation period bimodality, originally discovered by \citet{mcquillan2013}, is recovered for stars out to 525pc, but is not detectable at further distances. 
 We also find a significant width in the stellar main sequence of $M_G\sim$0.25 mag, as well as a  increase in the average rotation period correlated with the $M_G$ offset at a given color (mass).
-We interpret this to be the measurable change in luminosity and loss of angular momentum as stars evolve along the main sequence, which may provide a new independent test of stellar evolution and gyrochronlogy models.
-This investigation represents the first step in understanding the star formation history of our solar neighborhood as traced through stellar angular momentum loss. 
+We interpret this to be the measurable change in luminosity and loss of angular momentum as stars evolve along the main sequence, which may provide a new independent test of both stellar evolution and gyrochronlogy models.
+%This investigation represents the first step in understanding the star formation history of our solar neighborhood as traced through stellar angular momentum loss. 
 \end{abstract}
@@
 
-The \Kepler mission \citep{borucki2010} a new era for enabling detailed study of angular momentum loss in low-mass stars.
-long noted as a means to possibly age-date stars (gyro), open clusters give hope that this general model works for low-mass stars
-many Qs exist about details.
-these include Qs about initial rotation period distribution \citep[e.g.][]{barnes2010,matt2015}, the specific prescription for spin-down \citep{angus2015}, and exploring the efficiency of this angular momentum loss mechanism at old ages \citep{van-saders2016}.
+The \Kepler mission \citep{borucki2010}, through its unmatched combination of photometric precision and years-long observation baseline, has produced the largest sample of rotating stars to date.
+Stellar rotation has long been noted as a means to possibly age-date stars due to their constant angular momentum loss via winds
+\citep{skumanich1972}.
+While studies of open clusters give hope that this ``gyrochronology'' model broadly works for solar-type and lower-mass stars, many uncertainties exist about the details of this spin-down and its utility as a clock.
+These include questions about the initial rotation period distribution for stars \citep[e.g.][]{barnes2010,matt2015}, the specific analytic prescription for modeling the spin-down \citep{angus2015}, and exploring the efficiency of this angular momentum loss mechanism at older ages \citep{van-saders2016}.
 
-one of the most compelling results from the rotation work in \Kepler is the discovery of a period bimodality. 
-\citet{mcquillan2013} found bimodal feature, in M and later K stars \citep{mcquillan2014}.
-Davenport used Gaia DR1 to remove contamination from sub-giants and found the feature in G dwarfs.
-this feature proposed to be either a new short-lived transition or instability phase of momentum loss, or a signature of star formation history imprinted in the present-day rotation period distribution.
+One of the most compelling results from the rotating star sample in \Kepler is the discovery of a bimodal period distribution.
+\citet{mcquillan2013} first found a bimodality in the distribution of field M dwarfs with periods between $\sim$10 and $\sim$50 days.
+This feature was also found in the \Kepler field K dwarfs in \citep{mcquillan2014}.
+Using Gaia DR1, \citet{davenport2017} was able to remove contaminating sub-giants from the rotation sample and found the bimodality extended to the G dwarfs as well.
+This bimodal rotation period distribution is either a new short-lived transition or instability phase of angular momentum loss, or a signature of star formation history imprinted in the present-day rotation period distribution.
 However, to date this feature has only been observed in the \Kepler rotation period catalog, and most critically only for stars within $\sim$300 pc.
 
-In this paper we follow the work of \citet{davenport2017} in studying the \Kepler rotation period sample using astrometric data from the {\em Gaia} mission \citep{gaia}.
-By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2 \citep{gaia_dr2}, we can use precise distances for essentially every star to select likely main sequence dwarfs to distances well over 1 kpc.
-Importantly this filters out both sub-giants (the main contaminant noted by \cite{davenport2017}), and unresolved binary stars.
+In this paper we follow the work of \citet{davenport2017} in refining the \Kepler rotation period sample using astrometric data from the {\em Gaia} mission \citep{gaia}.
+By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2 \citep{gaia_dr2}, we can use precise distances for essentially every star to select the most-likely main sequence dwarfs to distances well over 1 kpc.
+Importantly this filters out both sub-giants, the main contaminant noted by \cite{davenport2017}, and unresolved binary stars.
 Here we demonstrate the power of such a combined time-domain and astrometric sample for constraining the detailed evolution of main sequence stars themselves, and exploring the star formation history of the Milky Way.
@@
 \centering
-\includegraphics[width=3.5in]{../figures/dist_Z}
+\includegraphics[width=3.3in]{../figures/dist_Z}
 \includegraphics[width=3.5in]{../figures/delta_per_subZ}
@@
 %\clearpage
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-A subtle feature that we noticed in Figure \ref{fig:cmd} is the apparent color gradient (i.e.
-rotation period gradient) between the single and binary star main sequence populations.
+A subtle feature we noticed in Figure \ref{fig:cmd} is the color gradient (i.e.
+rotation period gradient) for the stars in between the single and binary star main sequence populations.
 In Figure \ref{fig:cmd} this appears as a yellow stripe (i.e.

### 2018-06-25 paper/ms.tex
just have conclusion left

@@
 We have matched the astrometric data from {\em Gaia} Data Release 2 to the sample of stars with measured rotation periods from \Kepler.
-Using 30,305 stars with good distance estimates, we select 16,248 as being likely main sequence single stars centered within a 0.5 mag region about a 1 Gyr isochrone.
-This removes many sub-giants and unresolved binary stars from the sample.
+Using 30,305 stars with good distance estimates, we select 16,248 as being likely main sequence single stars centered within a 0.5 mag region about a 1 Gyr isochrone, removing many sub-giants and unresolved binary stars from the sample.
 The rotation period bimodality, originally discovered by \citet{mcquillan2013}, is recovered for stars out to 525pc, but is not detectable at further distances.
-We also find a significant width in the stellar main sequence of $M_G\sim$0.25 mag, as well as a  increase in the average rotation period correlated with the $M_G$ offset at a given color (mass).
+This bimodality correlates with Galactic height as well, dropping strongly for stars above $Z>90$ pc
+We also find a significant width in the stellar main sequence of $M_G\sim$0.25 mag, as well as an increase in the average rotation period correlated with the $M_G$ offset at a given color (mass).
 We interpret this to be the measurable change in luminosity and loss of angular momentum as stars evolve along the main sequence, which may provide a new independent test of both stellar evolution and gyrochronlogy models.
@@
 not a bimodal distance solution) and {\tt result\_flag} == 1 (i.e.
-well constrained distance).
+a well constrained distance).
 
-Our final sample contained 30,305 stars in Gaia DR2 with measured \Kepler rotation periods that passed these selection criteria.
-A color--magnitude diagram of this sample is presented in Figure \ref{fig:cmd}, with points colored by their rotation periods.
+Our final sample contained 30,305 stars in {\em Gaia} DR2 with measured \Kepler rotation periods that passed these selection criteria.
+A color--magnitude diagram (CMD) of this sample using the {\em Gaia} bands is presented in Figure \ref{fig:cmd}, with points colored by their measured \Kepler rotation periods.
 
@@
 \caption{
-Color--magnitude diagram for 30,305 \Kepler stars from the \citet{mcquillan2014} sample that are included in Gaia DR2, colored by their measured rotation period.
+Color--magnitude diagram for 30,305 \Kepler stars from the \citet{mcquillan2014} sample that are included in {\em Gaia} DR2, colored by their measured rotation period.
 For reference we show a $10^9$ year MIST isochrone (black line) used to select likely main sequence, single stars (black line).
@@
 
-\vspace{1in}
-%\clearpage
-
 %%%%%%%%%%%%%%%%%%%%%%
@@
 Left: Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with stars are colored by their measured \Kepler rotation periods from \citet{mcquillan2014}.
-MIST isochrones at ages of $10^8$, $10^9$, and $10^{10}$ yr are shown for comparison (blue, black, and red lines).
-The predicted evolution of a 0.7 M$_\odot$ star is highlighted (black triangles).
-%The main sequence shows significant scatter, as well as a gradient in rotation period with slower rotators preferentially above the main sequence.
+MIST isochrones at ages of $10^8$, $10^9$, and $10^{10}$ yr are shown for comparison (blue, black, and red lines). The predicted evolution of a 0.7 M$_\odot$ star is highlighted (black triangles). 
 Right: Difference in $M_G$ from the $10^9$ yr MIST isochrone as a function of rotation period (point color again indicates rotation period, as in Left panel).
@@
 absolute magnitude) from the $10^9$ year MIST isochrone.
-Slower rotating stars are apparently brighter at a given stellar color.
+Slower rotating stars are brighter at a given color.
 
 We interpret this observation as being due to the increase in luminosity of stars across their main sequence lives, coupled with the angular momentum loss that underpins gyrochronology.
-Indeed, the sequence of MIST isochrones shown in Figure \ref{fig:cmd_zoom} with ages of $10^8$, $10^9$, and $10^{10}$ years clearly predicts an increase in brightness of the main sequence over time.
+Indeed, the three MIST isochrones shown in Figure \ref{fig:cmd_zoom} with ages of $10^8$, $10^9$, and $10^{10}$ years clearly predicts a net increase in brightness of the main sequence over time.
+We also track the color--magnitude evolution of a 0.7 $M_\odot$ star over its lifetime in Figure \ref{fig:cmd_zoom}, which shows the ``hook'' the star takes over its main sequence life.
 
-indeed, 0.7solarmass star traced over its life in both panels doesnt quite make sense. 
-inverting CMD into precise ages w/ Gaia beyond scope of this work, but this figure suggests the MS is too wide given a reasonable age range.
-consider: younger stars, presumably more metal rich, should be redder, moving in the wrong direction to the affect we see here.
-more stellar masses and careful work needed...
-vertical spread in this parameter space is OK for a star, but maybe some color impact we not considering....
-hmm.
+Inverting the color--magnitude diagram into precise ages for main sequence stars using {\em Gaia} is beyond scope of this work.
+However, Figure \ref{fig:cmd_zoom} clearly demonstrates that age information is not only present in the CMD for main sequence stars, but it may also be a larger amplitude effect than previously assumed.
+For example: younger, presumably more metal rich stars, should have redder main sequence colors.
+This in turn should move the entire main sequence track to the right in the CMD.
+However, we find older stars (slower rotators) appearing on the right side of the main sequence, due to their increased luminosity with age.
+The main sequence location of a star is therefore not simply due to its mass and metallicity in the {\em Gaia} era, but its age must also be taken into account.
 
-however, looking at 0.7M period vs lum evolution, we see that the period seems to also drop off too rapidly compared to a model.
-combining the MM09 spin-down formulae for a 0.7M star, with the MIST isochrone evolution, we see  the period range should go out to 60--80 days to get the right amplitude offset.
-That implies that at a given age (or vertical offset), the stars are rotating {\it slightly faster} than expected.
-This is qualitatively similar to the broken spin-down at a critical Rossby number model suggested by \citet{van-saders2016}, though we cannot definitively confirm such intriguing rotation evolution from this initial investigation.
-However, more detailed studies of \Kepler and K2 rotating stars matched to {\em Gaia} may be an ideal ensemble dataset to test the \citet{van-saders2016} model against.
+In Figure \ref{fig:cmd_zoom} we also show the change in brightness as a function of the rotation period for the stars in our zoomed-in CMD, as well as a prediction for the evolution of a 0.7 $M_\odot$ star.
+The spin-down model here is a combination of the MIST isochrone evolution, as well as the \citet{meibom2009} gyrochronology model.
+Note this spin-down model has been arbitrarily offset in the $\Delta M_G$ to roughly go through the observed data at the rapid rotation end.
+While \Kepler light curves generally are not able to measure rotation periods reliably with periods longer than 20 or 30 days, we find that stars with brightness (vertical) offsets in the CMD consistent with being several Gyr old have average rotation periods far shorter than the 60--80 days values predicted from the spin-down models.
+
+One interpretation of this result may be that the older stars are spinning {\it faster} than expected for stars their age.
+This is qualitatively similar to the model of broken spin-down occurring at a critical Rossby number suggested by \citet{van-saders2016}.
+Though we cannot definitively confirm such intriguing rotation evolution from this initial investigation given the observation bias for long periods from \Kepler, matching the {\em Gaia} CMD with other rotation period measurements may provide an ideal dataset to test the \citet{van-saders2016} model against.
+The ability to simultaneously constrain the precise CMD location and surface rotation rate, as with the sample demonstrated here, may also be a valuable method for calibrating a new generation of isochrone and gyrochronology models.
 
@@
 \section{Discussion}
+%% HERE
 
@@
 
-This project was developed as part of the 2018 NYC Gaia Sprint, hosted by the Center for Computational Astrophysics at the Simons Foundation in New York City.
+This project was developed in part at the 2018 Gaia Sprint, hosted by the eScience and DIRAC Institutes at the University of Washington, Seattle, in parallel with the 2018 NYC Gaia Sprint, hosted by the Center for Computational Astrophysics at the Simons Foundation in New York City.
 

### 2018-06-28 paper/ms.tex
Wed progress

@@
 
+Using this sample of likely single, main sequence stars from \Kepler and {\em Gaia}, we are able to explore the distribution of rotation periods for stars as a function of their distance.
+The rotation period bimodality in \Kepler stars previously only detected for stars within $\sim$300 pc of the Sun due to the limits of available parallax data.
+Now with {\em Gaia} DR2, our sample of main sequence star with measured rotation periods in \Kepler extends out to over 2 kpc.
+
+In Figure \ref{fig:color_period} we present the period--color diagram for our sample of stars, split into six bins of projected distance.
+The first panel (0--350 pc) effectively reproduces the results of \citet{davenport2017} for bluer stars and \citet{mcquillan2014} for the redder stars.
+A gap in the observed rotation periods as a function of color is seen, at a period of approximately 5 days for $G_{BP}-G_{RP}\approx1$, 20 days for $G_{BP}-G_{RP}\approx2$, and increasing towards 30 days for the reddest stars in our sample.
+This gap corresponds with a line of approximately constant age, consistent with a gyrochrone with age $\sim$600 Myr \citep{davenport2017}.
+
 \begin{figure*}[]
@@
 
-Using this sample of likely single, main sequence stars from \Kepler and {\em Gaia}, we are able to explore the distribution of rotation periods for stars as a function of their distance.
-The rotation period bimodality in \Kepler stars previously only detected for stars within $\sim$300 pc of the Sun due to the limits of available parallax data.
-Now with {\em Gaia} DR2, our sample of main sequence star with measured rotation periods in \Kepler extends out to over 2 kpc.
-
-In Figure \ref{fig:color_period} we present the period--color diagram for our sample of stars, split into six bins of projected distance.
-The first panel (0--350 pc) effectively reproduces the results of \citet{davenport2017} for bluer stars and \citet{mcquillan2014} for the redder stars.
-A gap in the observed rotation periods as a function of color is seen, at a period of approximately 5 days for $G_{BP}-G_{RP}\approx1$, 20 days for $G_{BP}-G_{RP}\approx2$, and increasing towards 30 days for the reddest stars in our sample.
-This gap corresponds with a line of approximately constant age, consistent with a gyrochrone with age $\sim$600 Myr \citep{davenport2017}.
-
-The bimodality is still apparent in the second distance bin (350--525 pc), clearly visible in the redder stars, but seems to fade in the final three bins.
-Other structures in the period distributions are visible, however.
-For example, a thin sequence of stars with rotation periods near 10 days is faintly visible in the most distant bin (900--2500 pc) for stars with colors of $0.8<G_{BP}-G_{RP}<1.2$.
-This feature is due to the 1 Gyr open cluster NGC 6811 in the \Kepler field \citep{meibom2011}, whose distance is $\sim$1100 pc \citep{sandquist2016}.
-
-\begin{figure*}[!t]
+\begin{figure*}[!ht]
 \centering
@@
 \end{figure*}
+
+The bimodality is still apparent in the second distance bin (350--525 pc), clearly visible in the redder stars, but seems to fade in the final three bins.
+Other structures in the period distributions are visible, however.
+For example, a thin sequence of stars with rotation periods near 10 days is faintly visible in the most distant bin (900--2500 pc) for stars with colors of $0.8<G_{BP}-G_{RP}<1.2$.
+This feature is due to the 1 Gyr open cluster NGC 6811 in the \Kepler field \citep{meibom2011}, whose distance is $\sim$1100 pc \citep{sandquist2016}.
 
@@
 \caption{
-Top: projected distance and Galactic height of the main sequence stars in our sample.
-Two approximately equal subsets of stars near the Galactic mid-plane have been highlighted, those with distances from the sun of $d<300$ pc (red), and with $300<d<525$ pc.
+Top: Galactic height versus distance for the stars in our sample.
+Two approximately equal subsets of stars near the Galactic mid-plane have been highlighted, with distances of $d<300$ pc (red), and $300<d<525$ pc (blue).
 Bottom: As in Figure \ref{fig:per_hist}, the distribution of rotation periods after a 600 Myr gyrochrone was subtracted (faint lines), and with two-Gaussian fits (bold lines) for the same subsets of stars as above.
@@
 
+%\clearpage
 %%%%%%%%%%%%%%%%%%%%%%
@@
 Inverting the color--magnitude diagram into precise ages for main sequence stars using {\em Gaia} is beyond scope of this work.
-However, Figure \ref{fig:cmd_zoom} clearly demonstrates that age information is not only present in the CMD for main sequence stars, but it may also be a larger amplitude effect than previously assumed.
+However, Figure \ref{fig:cmd_zoom} clearly demonstrates that age information is not only present for main sequence stars, but it may also be a critical factor in accurately modeling the CMD.
 For example: younger, presumably more metal rich stars, should have redder main sequence colors.
@@
 \section{Discussion}
+
+Using a sample of 16,248 single main sequence stars with measured rotation periods from \Kepler and parallaxes from {\em Gaia} DR2, we have begun to explore the spatial distribution of stellar ages near the Sun using gyrochronology.
+The bimodality in rotation periods first reported by \citet{mcquillan2013} appear to be constrained to low galactic heights, rather than in a bubble centered around the Sun.
+This is consistent with the rotation period bimodality being a direct tracer of the star formation history, and indicates a burst of star formation within the past $\sim$600 Myr within the galactic disk.
+
+Unfortunately the \Kepler survey only covers a single pointing, and so our volume analysis of the stellar ages is limited here.
+However, with more than 15 distinct lines of sight available from K2, the extended \Kepler mission, we will soon be able to more than double our sample of rotating field stars and probe a much wider range of stellar ages.
+This may enable us to determine the spatial scales that star formation histories are coherant over in the nearby Milky Way, and make comparisons to the $\sim$100 pc resolved star formation history maps available for nearby galaxies such as Andromeda \citep[][]{lewis2015,williams2017}.
+For example, gyrochronology ages for stars in disk within 1 kpc of the Sun could be used to test if star formation propagates due to spiral arm density waves \citep[e.g.
+see][]{choi2015}.
+
+While we have removed as contaminants the prominent parallel main sequence of nearly equal-mass binary stars in this analysis, as well as sub-giant stars, both of these samples are clearly deserving of further analysis.
 %% HERE
-
-% from paper1
-%
-%The period bimodality may yet be a manifestation of the ``Vaughan-Preston'' gap observed in chromospheric activity indicators from solar type stars.
-Such a feature has also been discussed for rotating stars by \citet{kado-fong2016}.
-Given that the mass range for the bimodality explored
-
-we find period bimodality is localized to low galactic heights, not only close distances to the Sun.
-using rotation to trace the ages of nearby stars. would be really neat to compare the spatial structure of star formation in detail, as 
-\citet{lewis2015} for example show lots of variation in SFH in 100pc bins for M31
-%\citet{williams2017} also for M31, older SFH tracers
-and find out what the actual extent/coherance of the bimodality is.
-
-Adding to the sample of rotation periods from Kepler using K2 (Foreman-Mackey in prep?), not only for doubling the sample size to investigate trends with more precision, but also the wider range of galactic fields of view mean wider range of ages/SFH's we can sample.
-
-We note a shortcoming of our approach in normalizing the rotation period distributions to a 600 Myr gyrochrone in Figure \ref{fig:per_hist} is that it gives the false appearance that this age is a universal transition point.
-A more robust approach is needed that uses improved gyrochronology models to invert the observed color--period diagrams into age distributions.
 
@@
 
+\software{Python, IPython \citep{ipython}, NumPy \citep{numpy}, Matplotlib \citep{matplotlib}, SciPy \citep{scipy}, Pandas \citep{pandas}, Astropy \citep{astropy}}
+
 \bibliography{/Users/james/Dropbox/references}

### 2018-06-29 paper/ms.tex
paper full draft finished

@@
 
+\clearpage
 %%%%%%%%%%%%%%%%%%%%%%
@@
 This may enable us to determine the spatial scales that star formation histories are coherant over in the nearby Milky Way, and make comparisons to the $\sim$100 pc resolved star formation history maps available for nearby galaxies such as Andromeda \citep[][]{lewis2015,williams2017}.
-For example, gyrochronology ages for stars in disk within 1 kpc of the Sun could be used to test if star formation propagates due to spiral arm density waves \citep[e.g.
+For example, gyrochronology ages for disk stars within 1 kpc of the Sun could be used to test if star formation propagates due to spiral arm density waves \citep[e.g.
 see][]{choi2015}.
@@
 While we have removed as contaminants the prominent parallel main sequence of nearly equal-mass binary stars in this analysis, as well as sub-giant stars, both of these samples are clearly deserving of further analysis.
-%% HERE
+For example, comparing the rotation period distribution for binary versus single main sequence stars may provide constraints on their dynamical histories.
 
-finally, this work has found strong motivation for a new generation of gyrochronology isochrones to adequately model the {\em Gaia}--\Kepler/K2 (and soon TESS) rotation data for field stars.
-As we show in Figure \ref{fig:cmd_zoom}, these models should include both the angular momentum evolution (period), as well as the temperature and luminosity (color, magnitude) as a function of time.
-With such improvements, we may be able to
+Finally, in the {\em Gaia} era, stellar evolution modelers have the daunting task to explain an increasing number of precisely determined features from survey observations, such as the fascinating new main sequence gap revealed by \citet{jao2018}, and the main sequence width seen here in Figure \ref{fig:cmd}.
+This work has presented strong motivation for a new generation of stellar evolution and rotation models to be produced to accurately model the {\em Gaia} main sequence.
+With \Kepler, K2, and soon TESS, rotation measurements will be available for hundreds of thousands of nearby stars that {\em Gaia} has provided precise distances for.
+We hope new isochrone models will reproduce both the observed colors and luminosities of stars due to their ages and abundances, as well as the evolution of their surface rotations.
 

### 2018-07-19 paper/ms.tex
paper edited from KRC

@@
 Using 30,305 stars with good distance estimates, we select 16,248 as being likely main sequence single stars centered within a 0.5 mag region about a 1 Gyr isochrone, removing many sub-giants and unresolved binary stars from the sample.
-The rotation period bimodality, originally discovered by \citet{mcquillan2013}, is recovered for stars out to 525pc, but is not detectable at further distances.
-This bimodality correlates with Galactic height as well, dropping strongly for stars above $Z>90$ pc
-We also find a significant width in the stellar main sequence of $M_G\sim$0.25 mag, as well as an increase in the average rotation period correlated with the $M_G$ offset at a given color (mass).
-We interpret this to be the measurable change in luminosity and loss of angular momentum as stars evolve along the main sequence, which may provide a new independent test of both stellar evolution and gyrochronlogy models.
-%This investigation represents the first step in understanding the star formation history of our solar neighborhood as traced through stellar angular momentum loss. 
+The rotation period bimodality, originally discovered by \citet{mcquillan2013}, is clearly recovered for stars out to 525pc, but is not detectable at further distances.
+This bimodality correlates with Galactic height as well, dropping strongly for stars above $Z>90$ pc.
+We also find a significant width in the stellar main sequence of $M_G\sim$0.25 mag, as well as a coherent gradient of increasing rotation periods diagonal to the main sequence.
+We interpret this to be the simultaneous change in luminosity and loss of angular momentum as stars evolve along the main sequence.
+The diagonal offset from the main sequence is unexpected, however, and may provide a new independent test of both stellar evolution and gyrochronlogy models.
 \end{abstract}
@@
 \section{Introduction}
+Over the past decade the study of stellar rotation has transformed from a niche observation limited to a handful of nearby clusters and few thousand bright stars, to an area of great  interest for field stars with a rapidly growing sample size.
+The \Kepler mission \citep{borucki2010}, through its unmatched combination of photometric precision and years-long observation baseline, has produced the largest precision sample of rotating stars to date -- more than 30,000 sources to date \citep{mcquillan2014}.
+Although the light curves available from the {\em Gaia} mission are not as densely sampled in time as \Kepler \citep{gaia}, they have yielded over 147,000 candidate rotation periods as of Data Release 2 \citep{lanzafame2018}.
+The extended \Kepler mission, K2 \citep{howell2014}, has now observed more stars than the original \Kepler sample, potentially doubling the number of high precision stellar rotation periods available.
+Similarly, by the end of the {\em Gaia} mission \citet{lanzafame2018} estimate a sample of 3--20 million rotation periods will be recovered.
+The era of statistical studies in stellar rotation has therefore begun.
 
-The \Kepler mission \citep{borucki2010}, through its unmatched combination of photometric precision and years-long observation baseline, has produced the largest sample of rotating stars to date.
-Stellar rotation has long been noted as a means to possibly age-date stars due to their constant angular momentum loss via winds
-\citep{skumanich1972}.
+Stellar rotation has long been noted as a means to possibly age-date stars due to their constant angular momentum loss via winds \citep{skumanich1972}. 
 While studies of open clusters give hope that this ``gyrochronology'' model broadly works for solar-type and lower-mass stars, many uncertainties exist about the details of this spin-down and its utility as a clock.
@@
 \citet{mcquillan2013} first found a bimodality in the distribution of field M dwarfs with periods between $\sim$10 and $\sim$50 days.
-This feature was also found in the \Kepler field K dwarfs in \citep{mcquillan2014}.
-Using Gaia DR1, \citet{davenport2017} was able to remove contaminating sub-giants from the rotation sample and found the bimodality extended to the G dwarfs as well.
-This bimodal rotation period distribution is either a new short-lived transition or instability phase of angular momentum loss, or a signature of star formation history imprinted in the present-day rotation period distribution.
-However, to date this feature has only been observed in the \Kepler rotation period catalog, and most critically only for stars within $\sim$300 pc.
+This feature was also found in the \Kepler field K dwarfs in \citet{mcquillan2014}, but was not seen in the bluer stars.
+Using Gaia DR1, \citet{davenport2017} was able to remove contaminating sub-giants from the rotation sample, and found the bimodality extended to the G dwarfs as well.
+This bimodal surface rotation period distribution is either a new short-lived transition or instability phase of rapid angular momentum loss, or a signature of star formation history imprinted in the present-day rotation period distribution.
 
-In this paper we follow the work of \citet{davenport2017} in refining the \Kepler rotation period sample using astrometric data from the {\em Gaia} mission \citep{gaia}.
-By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2 \citep{gaia_dr2}, we can use precise distances for essentially every star to select the most-likely main sequence dwarfs to distances well over 1 kpc.
+However, as \citet{davenport2017} notes this feature has only been observed in the \Kepler rotation period catalog, and most critically only for stars within $\sim$300 pc.
+The faint M dwarfs in the \Kepler sample studied in \citet{mcquillan2013} have an average distance of $<$250 pc.
+Gaia DR1 was also only able to provide precise distances for the nearest stars, and \citet{davenport2017} report an median distance of 285 pc for the blue stars in their sample.
+
+In this paper we extend the work of \citet{davenport2017} in studying the \Kepler rotation period distribution as a function of distance from the Sun using new astrometric data from the {\em Gaia} mission.
+By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2 \citep{gaia_dr2}, we can use precise distances for essentially every star in the McQuillan catalog to select the most-likely main sequence dwarfs out to distances $>$2 kpc.
 Importantly this filters out both sub-giants, the main contaminant noted by \cite{davenport2017}, and unresolved binary stars.
@@
 The sample from \citet{mcquillan2014} provides rotation periods for more 34,030 stars, measured using the Auto-Correlation Function (ACF).
-While the ACF does not have as fine of resolution in recovering periods as compared with methods such as the Lomb-Scargle Periodogram, it is more robust to detecting the true period as opposed to an alias, and more complete for batch analysis of all stars \citep[e.g.
+While the ACF does not recover periods with as much precision as methods such as the Lomb-Scargle Periodogram, it is more robust to detecting the true period as opposed to an alias, and more complete for batch analysis of all stars \citep[e.g.
 see][]{aigrain2015}.
@@
 Bedell, which included entries for 195,830 sources.
-\Kepler-based stellar parameters included in this cross-match come the Data Release 25 \Kepler catalog.
-Joining this cross-matched table to the \citet{mcquillan2014} catalog, we found 33,538 sources.
+\Kepler-based stellar parameters are included in this cross-match from the Data Release 25 \Kepler catalog.
+Joining this cross-matched table to the \citet{mcquillan2014} catalog, we found 33,538 sources with {\em Gaia} astrometry and \Kepler-derived rotation periods.
 
@@
 
-Rather than simply use the inverse {\em Gaia} parallax values to measure the distance to sources, we use the improved distance prescription from \citet{bailer-jones2018}, who provided independent distances estimates for 1.33 billion {\em Gaia} sources using a weak prior on the distribution of stars in our galaxy.
-We follow their suggested use of the distance catalog, we only include sources with {\tt modality\_flag} == 1 (i.e.
+Rather than simply use the inverse {\em Gaia} parallax values to measure the distance to sources, we use the improved distance prescription from \citet{bailer-jones2018}, who provided independent distance estimates for 1.33 billion {\em Gaia} sources using a weak prior on the distribution of stars in our Galaxy.
+We follow their suggested use of the distance catalog, including only sources with {\tt modality\_flag} == 1 (i.e.
 not a bimodal distance solution) and {\tt result\_flag} == 1 (i.e.
@@
 Color--magnitude diagram for 30,305 \Kepler stars from the \citet{mcquillan2014} sample that are included in {\em Gaia} DR2, colored by their measured rotation period.
-For reference we show a $10^9$ year MIST isochrone (black line) used to select likely main sequence, single stars (black line).
-A track of binary stars is apparent $\sim$0.75 mag above the main sequence.
-As in \citet{davenport2017}, we find significant contamination of the rotation period sample for bluer stars by sub-giants.}
+For reference we show a $10^9$ year MIST isochrone used to select likely main sequence, single stars (black line).
+As in \citet{davenport2017}, the bluer stars in the sample are significantly contaminated by sub-giants.
+A track of binary stars is apparent $\sim$0.75 mag above the main sequence. The inset figure highlights a portion of the single star and equal-mass binary main sequences.
+}
 \label{fig:cmd}
@@
 As in \citet{davenport2017}, the color--magnitude diagram in Figure \ref{fig:cmd} shows many of the bluer stars in the \citet{mcquillan2014} sample are located significantly above the main sequence.
-These are likely subgiant stars, which do not follow the main sequence stars spin-down evolution \citep[e.g.][]{donascimento2012, van-saders2013}.
-Since \citet{davenport2017} found subgiants could obscure the rotation period bimodality for G dwarfs, these must be excluded from our analysis, but we encourage future studies to explore the wealth of angular momentum evolution data from these most-main sequence objects.
+These are likely subgiant stars, which no longer follow the main sequence stars spin-down evolution \citep[e.g.][]{donascimento2012, van-saders2013}.
+Since \citet{davenport2017} found subgiants could obscure the rotation period bimodality for G dwarfs, these must be excluded from our analysis, but we encourage future studies to explore the wealth of angular momentum evolution data from these post-main sequence objects.
 
-Beyond the subgiant contamination, we also see a secondary population of stars in a parallel track $\sim$0.75 mag above the normal main sequence, which are attributed to unresolved binary star systems.
-This pile-up above the main sequence occurs due to unresolved equal-mass (or nearly equal-mass) field binaries, and was seen in the {\em Gaia} DR1 data as well \citep{anderson2017}.
-Since these systems may have experienced tidal evolution that could significantly impact their rotation evolution \citep[e.g.][]{lurie2017}, we must also remove these from our analysis.
-Though we do not explore the binary population in any detail here, this sample (perhaps with radial velocity follow-up) may provide useful insight into the tidal evolution of binary stars, and are good targets for characterizing binary star system properties.
+Beyond the subgiant contamination, we also see a secondary population of stars in a parallel track $\sim$0.75 mag above the normal main sequence, as expected and seen in other color--magnitude diagrams.
+This parallel main sequence occurs due to unresolved equal-mass (or nearly equal-mass) field binaries, and was seen in the {\em Gaia} DR1 data as well \citep{anderson2017}.
+Since the tightest of these systems may have experienced tidal evolution that could significantly impact their rotation evolution \citep[e.g.][]{lurie2017}, and we lack the ability with the {\em Gaia} DR2 data to adequately constrain their physical separations,
+we must also remove these unresolved binaries from our analysis.
+We do not explore the binary population in any detail here, but this sample could provide useful insight into the tidal evolution of binary stars, and are good targets for radial velocity follow-up to characterize binary system properties.
 We also note a small number of systems above the even the equal-mass binary main sequence track, which could be due to unresolved triple star systems.
@@
 
-Using this sample of likely single, main sequence stars from \Kepler and {\em Gaia}, we are able to explore the distribution of rotation periods for stars as a function of their distance.
-The rotation period bimodality in \Kepler stars previously only detected for stars within $\sim$300 pc of the Sun due to the limits of available parallax data.
-Now with {\em Gaia} DR2, our sample of main sequence star with measured rotation periods in \Kepler extends out to over 2 kpc.
+Using this sample of likely single, main sequence stars from \Kepler and {\em Gaia}, we are able to explore the distribution of rotation periods for stars as a function of their distance and Galactic location.
+The rotation period bimodality in \Kepler stars was previously detected only for stars within $\sim$300 pc of the Sun due to the limits of available parallax data.
+Now with {\em Gaia} DR2, our sample of main sequence stars with measured rotation periods from \Kepler with adequate distance estimates for filtering out sub-giants extends to over 2 kpc.
 
@@
 \citet{davenport2017} used an ad-hoc modification of the gyrochrone model for the hottest stars to illustrate the existence of the period bimodality.
-However, in Figure \ref{fig:per_hist} we limit our analysis to stars with $0.8<B-V<1.5$ (approximately $0.9<G_{BP}-G_{RP}<2.2$, or $0.9>M_\odot> 0.5$), where the gyrochrone is most flat as a function of color.
+However, in Figure \ref{fig:per_hist} we limit our analysis to stars with $0.8<B-V<1.5$ (approximately $0.9<G_{BP}-G_{RP}<2.2$, or $0.5<M_\odot< 0.9$), where the gyrochrone is most flat as a function of color.
 This color range was chosen to ensure ample stars were available in each distance bin, but without having to use the ad-hoc gyrochrone model correction from \citet{davenport2017}.
 
-Figure \ref{fig:per_hist} (left) shows clearly that for the short period component of the rotation period bimodality dominates only for the nearest stars, and steadily decreases with increasing distance.
-%but can also cut vs Z instead of d. See bimodality drops neatly with height in Figure \ref{fig:per_hist} also.
+Figure \ref{fig:per_hist} (left) shows clearly that the short period component of the rotation period bimodality dominates only for the nearest stars, and steadily decreases with increasing distance.
 The \Kepler field is centered on a Galactic latitude of $b\sim13.5^\circ$, and so we can also study the period bimodality as a function of height above the Galactic mid-plane.
-Both galaxy formation simulations \citep{ma2017} and observations of stars in the nearby Milky Way \citep{xiang2017} indicate that height above the mid-plane correlates strongly with the median ages for stars, out to distances of several kpc.
-The Milky Way thin disk has a scaleheight of $\sim$300 pc \citep{gilmore1983}.
+Both galaxy formation simulations \citep{ma2017} and observations of stars in the nearby Milky Way \citep{xiang2017} indicate that height above the mid-plane correlates strongly with the median ages for stars out to distances of several kpc.
+For low-mass stars the Milky Way thin disk has a scaleheight of $\sim$300 pc \citep{gilmore1983}.
 However, since the \Kepler field is oriented towards low latitudes we do not reach significant {\it heights} above the disk.
@@
 
-As a first attempt at understanding the spatial extent of this age-related feature, in Figure \ref{fig:dZ} we break the lowest height stars ($Z < 100$ pc) in to two roughly even samples, split as a function of their projected distance.
+To further understanding the spatial extent of this age-related feature, in Figure \ref{fig:dZ} we break the lowest height stars ($Z < 100$ pc) in to two roughly even samples, split as a function of their projected distance.
 Note we have also repeated this exercise for stars in higher ranges of $Z$, and find the decline of the rapid rotators is again uniform between subsamples of varying projected distance.
@@
 \centering
-\includegraphics[width=3.3in]{../figures/dist_Z}
+\includegraphics[width=3.4in]{../figures/dist_Z}
 \includegraphics[width=3.5in]{../figures/delta_per_subZ}
@@
 \caption{
-Left: Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with stars are colored by their measured \Kepler rotation periods from \citet{mcquillan2014}.
+Left: Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with main sequence stars colored by their measured rotation periods from \citet{mcquillan2014}.
 MIST isochrones at ages of $10^8$, $10^9$, and $10^{10}$ yr are shown for comparison (blue, black, and red lines). The predicted evolution of a 0.7 M$_\odot$ star is highlighted (black triangles). 
@@
 We interpret this observation as being due to the increase in luminosity of stars across their main sequence lives, coupled with the angular momentum loss that underpins gyrochronology.
-Indeed, the three MIST isochrones shown in Figure \ref{fig:cmd_zoom} with ages of $10^8$, $10^9$, and $10^{10}$ years clearly predicts a net increase in brightness of the main sequence over time.
+Indeed, the three MIST isochrones shown in Figure \ref{fig:cmd_zoom} with ages of $10^8$, $10^9$, and $10^{10}$ years clearly predict a net increase in brightness of the main sequence over time.
 We also track the color--magnitude evolution of a 0.7 $M_\odot$ star over its lifetime in Figure \ref{fig:cmd_zoom}, which shows the ``hook'' the star takes over its main sequence life.
@@
 
-In Figure \ref{fig:cmd_zoom} we also show the change in brightness as a function of the rotation period for the stars in our zoomed-in CMD, as well as a prediction for the evolution of a 0.7 $M_\odot$ star.
+In Figure \ref{fig:cmd_zoom} we also show each stars' offset from the main sequence as a function of its rotation period within our zoomed-in CMD, as well as a prediction for the evolution of a 0.7 $M_\odot$ star.
 The spin-down model here is a combination of the MIST isochrone evolution, as well as the \citet{meibom2009} gyrochronology model.
-Note this spin-down model has been arbitrarily offset in the $\Delta M_G$ to roughly go through the observed data at the rapid rotation end.
-While \Kepler light curves generally are not able to measure rotation periods reliably with periods longer than 20 or 30 days, we find that stars with brightness (vertical) offsets in the CMD consistent with being several Gyr old have average rotation periods far shorter than the 60--80 days values predicted from the spin-down models.
+Note this spin-down model has been arbitrarily offset in $\Delta M_G$ to approximately match the observed rapid rotators.
+While \Kepler light curves are not able to measure rotation reliably for stars with periods longer than $\sim$30 days, we find that stars with brightness (vertical) offsets in the CMD consistent with being several Gyr old have average rotation periods far shorter than the 60--80 days values predicted from the spin-down models.
 
@@
 This is qualitatively similar to the model of broken spin-down occurring at a critical Rossby number suggested by \citet{van-saders2016}.
-Though we cannot definitively confirm such intriguing rotation evolution from this initial investigation given the observation bias for long periods from \Kepler, matching the {\em Gaia} CMD with other rotation period measurements may provide an ideal dataset to test the \citet{van-saders2016} model against.
-The ability to simultaneously constrain the precise CMD location and surface rotation rate, as with the sample demonstrated here, may also be a valuable method for calibrating a new generation of isochrone and gyrochronology models.
+Though we cannot definitively confirm such intriguing rotation evolution from this initial investigation given the observation bias for long rotation periods from \Kepler, matching the {\em Gaia} CMD with other rotation period measurements may provide an ideal dataset to test the \citet{van-saders2016} model against.
 
-\clearpage
+In tracking only the brightness evolution of stars as they age, we neglect the color evolution shown in Figure \ref{fig:cmd_zoom} (left panel), which predicts a star will move roughly {\it along} the main sequence track as it changes in luminosity over time.
+Instead we clearly observe a diagonal gradient in the rotation periods consistent with being a stellar age affect, which may point to missing physics in the MIST models.
+We believe simultaneous constraints on both the precise CMD location and surface rotation rate for stars, as demonstrated here, will be a valuable utility for calibrating a new generation of isochrone and gyrochronology models.
+
+%\clearpage
 %%%%%%%%%%%%%%%%%%%%%%

### 2018-07-20 paper/ms.tex
comments round 2 from KRC

@@
 This bimodality correlates with Galactic height as well, dropping strongly for stars above $Z>90$ pc.
-We also find a significant width in the stellar main sequence of $M_G\sim$0.25 mag, as well as a coherent gradient of increasing rotation periods diagonal to the main sequence.
+We also find a significant width in the stellar main sequence of $\Delta M_G\sim$0.25 mag, as well as a coherent gradient of increasing rotation periods diagonal to the main sequence.
 We interpret this to be the simultaneous change in luminosity and loss of angular momentum as stars evolve along the main sequence.
@@
 
-%\clearpage
 %%%%%%%%%%%%%%%%%%%%%%
@@
 \centering
-\includegraphics[width=6in]{../figures/cmd_zoom}
+\includegraphics[width=7in]{../figures/cmd_zoom}
 \caption{
-Left: Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with main sequence stars colored by their measured rotation periods from \citet{mcquillan2014}.
+{\bf Left:} Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with main sequence stars colored by their measured rotation periods from \citet{mcquillan2014}.
 MIST isochrones at ages of $10^8$, $10^9$, and $10^{10}$ yr are shown for comparison (blue, black, and red lines). The predicted evolution of a 0.7 M$_\odot$ star is highlighted (black triangles). 
-Right: Difference in $M_G$ from the $10^9$ yr MIST isochrone as a function of rotation period (point color again indicates rotation period, as in Left panel).
-Blue squares show an increase in the median $M_G$ offset in bins of increasing rotation period.
-Error bars shown are the standard deviation of $\Delta M_G$ in each bin.
-The 0.7 M$_\odot$ star brightness evolution is shown as a function of rotation evolution from \citet{meibom2009} (black triangles).
+{\bf Center:} Difference in $M_G$ from the $10^9$ yr MIST isochrone as a function of rotation period (point color again indicates rotation period, as in Left panel).
+An increase in the median brightness is seen in bins of rotation period (blue squares).
+Error bars are the standard deviation of $\Delta M_G$ in each bin.
+The 0.7 M$_\odot$ star brightness evolution from MIST is shown as a function of rotation evolution from \citet{meibom2009}, manually shifted to match the rapid rotating stars (black triangles).
+{\bf Right:} Same as the center panel, but representing the change in $G_{BP}-G_{RP}$ color as a function of measured rotation period.
+Again the MIST--Meibom model evolution (black triangles) has been shifted to match the rapid rotators.
+Stars are predicted to get {\it bluer} at older ages, rather than redder as our sample indicates.
 }
@@
 
-A subtle feature we noticed in Figure \ref{fig:cmd} is the color gradient (i.e.
+A subtle feature we noticed in Figure \ref{fig:cmd} is the {\it diagonal} color gradient (i.e.
 rotation period gradient) for the stars in between the single and binary star main sequence populations.
@@
 To exaggerate this feature, we have reproduced a portion of our color--magnitude diagram focused on the main sequence near this stellar color in the left panel of Figure \ref{fig:cmd_zoom}.
-A clear color gradient is present, with red points (slower rotators) appearing preferentially above the main sequence.
-The right panel of Figure \ref{fig:cmd_zoom} concretely demonstrates a correlation between the measured rotation period and the vertical offset (i.e.
+A clear color gradient is present, with red points (slower rotators) appearing preferentially above and to the right of the main sequence.
+The center panel of Figure \ref{fig:cmd_zoom} demonstrates a correlation between the measured rotation period and the vertical offset (i.e.
 absolute magnitude) from the $10^9$ year MIST isochrone.
 Slower rotating stars are brighter at a given color.
+However, as the right panel in Figure \ref{fig:cmd_zoom} shows, the MIST models predict a bluer color for stars at older ages, while our data show the slower rotators have redder colors on average.
 
-We interpret this observation as being due to the increase in luminosity of stars across their main sequence lives, coupled with the angular momentum loss that underpins gyrochronology.
-Indeed, the three MIST isochrones shown in Figure \ref{fig:cmd_zoom} with ages of $10^8$, $10^9$, and $10^{10}$ years clearly predict a net increase in brightness of the main sequence over time.
-We also track the color--magnitude evolution of a 0.7 $M_\odot$ star over its lifetime in Figure \ref{fig:cmd_zoom}, which shows the ``hook'' the star takes over its main sequence life.
+In the center and right panels of Figure \ref{fig:cmd_zoom} we show a prediction of the brightness and color evolution of a 0.7 $M_\odot$ star as a function of its rotation period over time.
+The predicted evolution here is a combination of the MIST isochrone models from $10^8$ to $10^{10}$ years, as well as the \citet{meibom2009} gyrochronology model over this same time window.
+Note this  combined model has been arbitrarily offset in $\Delta M_G$ and $\Delta(G_{BP} - G_{RP})$ in the middle and right panels, respectively, to approximately match the CMD position observed for the rapid rotators.
 
-Inverting the color--magnitude diagram into precise ages for main sequence stars using {\em Gaia} is beyond scope of this work.
-However, Figure \ref{fig:cmd_zoom} clearly demonstrates that age information is not only present for main sequence stars, but it may also be a critical factor in accurately modeling the CMD.
-For example: younger, presumably more metal rich stars, should have redder main sequence colors.
-This in turn should move the entire main sequence track to the right in the CMD.
-However, we find older stars (slower rotators) appearing on the right side of the main sequence, due to their increased luminosity with age.
-The main sequence location of a star is therefore not simply due to its mass and metallicity in the {\em Gaia} era, but its age must also be taken into account.
+These data challenge our main sequence model evolution in two ways.
+Firstly, the diagonal gradient from the main sequence in rotation period is in conflict with the evolution predicted from MIST.
+The stellar evolution model shows that a star should evolve essentially {\it along} the main sequence track, as in the left panel of Figure \ref{fig:cmd_zoom}.
+The tension with our observed sample is highlighted in the right panel of Figure \ref{fig:cmd_zoom}, where the predicted colors for old stars get bluer, while our slice through the observed rotation period sample shows older stars being redder.
 
-In Figure \ref{fig:cmd_zoom} we also show each stars' offset from the main sequence as a function of its rotation period within our zoomed-in CMD, as well as a prediction for the evolution of a 0.7 $M_\odot$ star.
-The spin-down model here is a combination of the MIST isochrone evolution, as well as the \citet{meibom2009} gyrochronology model.
-Note this spin-down model has been arbitrarily offset in $\Delta M_G$ to approximately match the observed rapid rotators.
-While \Kepler light curves are not able to measure rotation reliably for stars with periods longer than $\sim$30 days, we find that stars with brightness (vertical) offsets in the CMD consistent with being several Gyr old have average rotation periods far shorter than the 60--80 days values predicted from the spin-down models.
+This tension in the CMD evolution is compounded when we consider stellar metallicity, which we have not varied in our MIST model realization.
+Younger, presumably more metal rich stars should have redder optical colors on the main sequence, while older metal poor stars should be more blue.
+Instead we find that older, slower rotating stars are redder and brighter than their rapidly rotating counterparts.
+While we do not have independent metallicity constraints for the \Kepler stars in this sample, Figure \ref{fig:cmd_zoom} shows that age may have as big of a factor as metallicity in the precise CMD location of main sequence stars.
+Likewise, stellar evolution models do not appear to reproduce the CMD positions for main sequence, low-mass stars of all ages.
 
-One interpretation of this result may be that the older stars are spinning {\it faster} than expected for stars their age.
-This is qualitatively similar to the model of broken spin-down occurring at a critical Rossby number suggested by \citet{van-saders2016}.
-Though we cannot definitively confirm such intriguing rotation evolution from this initial investigation given the observation bias for long rotation periods from \Kepler, matching the {\em Gaia} CMD with other rotation period measurements may provide an ideal dataset to test the \citet{van-saders2016} model against.
+The second, more subtle challenge presented by these data is the rotation periods observed for presumably older stars. 
+We find that stars having $M_G$ offsets in the center panel of Figure \ref{fig:cmd_zoom} consistent with being several Gyr old have average rotation periods of 30--40 days, far shorter than the 60+ days values predicted by the spin-down model for a 0.7 $M_\odot$ star.
+Note, \Kepler light curves are often not able to reliably measure rotation periods longer than $\sim$30 days.
+This bias means we may be missing even slower rotators from our \Kepler--{\em Gaia} combined sample, and do not know what the $M_G$ offset for such stars would be.
 
-In tracking only the brightness evolution of stars as they age, we neglect the color evolution shown in Figure \ref{fig:cmd_zoom} (left panel), which predicts a star will move roughly {\it along} the main sequence track as it changes in luminosity over time.
-Instead we clearly observe a diagonal gradient in the rotation periods consistent with being a stellar age affect, which may point to missing physics in the MIST models.
-We believe simultaneous constraints on both the precise CMD location and surface rotation rate for stars, as demonstrated here, will be a valuable utility for calibrating a new generation of isochrone and gyrochronology models.
+One possible interpretation of this result as observed here is that the older stars are spinning {\it faster} than expected for stars their age.
+This is qualitatively similar to the model of broken spin-down occurring at a critical Rossby number suggested by \citet{van-saders2016}.Though we cannot definitively confirm such intriguing rotation evolution from this initial investigation given the observation bias for long rotation periods from \Kepler, matching the {\em Gaia} CMD with other rotation period measurements may provide an ideal dataset to test the \citet{van-saders2016} model against.
 
-%\clearpage
 %%%%%%%%%%%%%%%%%%%%%%
+\vspace{0.5in}
 \section{Discussion}
@@
 Using a sample of 16,248 single main sequence stars with measured rotation periods from \Kepler and parallaxes from {\em Gaia} DR2, we have begun to explore the spatial distribution of stellar ages near the Sun using gyrochronology.
-The bimodality in rotation periods first reported by \citet{mcquillan2013} appear to be constrained to low galactic heights, rather than in a bubble centered around the Sun.
-This is consistent with the rotation period bimodality being a direct tracer of the star formation history, and indicates a burst of star formation within the past $\sim$600 Myr within the galactic disk.
+The bimodality in rotation periods first reported by \citet{mcquillan2013} appear to be constrained to low Galactic scaleheights, rather than in an obvious bubble centered around the Sun.
+Since height above the Galactic plane is assumed to be related to age, this is consistent with the rotation period bimodality being a direct tracer of the star formation history, and indicates a burst of star formation within the past $\sim$600 Myr within the Galactic disk.
 
 Unfortunately the \Kepler survey only covers a single pointing, and so our volume analysis of the stellar ages is limited here.
-However, with more than 15 distinct lines of sight available from K2, the extended \Kepler mission, we will soon be able to more than double our sample of rotating field stars and probe a much wider range of stellar ages.
-This may enable us to determine the spatial scales that star formation histories are coherant over in the nearby Milky Way, and make comparisons to the $\sim$100 pc resolved star formation history maps available for nearby galaxies such as Andromeda \citep[][]{lewis2015,williams2017}.
+However, with 16 distinct lines of sight available from K2, the extended \Kepler mission, we will be able to more than double our sample of rotating field stars and probe a much wider range of stellar ages.
+This may enable us to determine the spatial scales over which star formation histories are coherant in the Milky Way disk.
+We could then make comparisons to the $\sim$100 pc resolved star formation history maps available for nearby galaxies such as Andromeda \citep[][]{lewis2015,williams2017}.
 For example, gyrochronology ages for disk stars within 1 kpc of the Sun could be used to test if star formation propagates due to spiral arm density waves \citep[e.g.
@@
 
-Finally, in the {\em Gaia} era, stellar evolution modelers have the daunting task to explain an increasing number of precisely determined features from survey observations, such as the fascinating new main sequence gap revealed by \citet{jao2018}, and the main sequence width seen here in Figure \ref{fig:cmd}.
-This work has presented strong motivation for a new generation of stellar evolution and rotation models to be produced to accurately model the {\em Gaia} main sequence.
-With \Kepler, K2, and soon TESS, rotation measurements will be available for hundreds of thousands of nearby stars that {\em Gaia} has provided precise distances for.
-We hope new isochrone models will reproduce both the observed colors and luminosities of stars due to their ages and abundances, as well as the evolution of their surface rotations.
+Finally, in the {\em Gaia} era, stellar evolution modelers have the daunting task of explaining an increasing number of precisely determined features in the CMD.
+This includes the fascinating new main sequence gap revealed by \citet{jao2018}, as well as the diagonal age (or rotation) gradient in the main sequence shown in our Figures \ref{fig:cmd} and \ref{fig:cmd_zoom}.
+Our work presents strong motivation for a new generation of stellar evolution and rotation models to accurately reproduce the main sequence as observed with \Kepler and {\em Gaia}.
+With \Kepler, K2, and soon TESS, precise rotation measurements will be available for hundreds of thousands of nearby stars that {\em Gaia} has provided reliable distances for.
+We hope these data will guide new isochrone models in reproducing both the observed colors and luminosities of stars due to their ages and abundances, as well as the evolution of their surface rotation rates.
 

### 2018-07-21 paper/ms.tex
submitted to AAS journals

@@
 This bimodality correlates with Galactic height as well, dropping strongly for stars above $Z>90$ pc.
-We also find a significant width in the stellar main sequence of $\Delta M_G\sim$0.25 mag, as well as a coherent gradient of increasing rotation periods diagonal to the main sequence.
-We interpret this to be the simultaneous change in luminosity and loss of angular momentum as stars evolve along the main sequence.
-The diagonal offset from the main sequence is unexpected, however, and may provide a new independent test of both stellar evolution and gyrochronlogy models.
+We also find a significant width in the stellar main sequence of $\Delta M_G\sim$0.25 mag, as well as a coherent gradient of increasing rotation periods orthogonal to the main sequence.
+We interpret this as a signature of stellar angular momentum loss over time, implying a corresponding diagonal age gradient across the main sequence.
+Stellar evolution models predict changes in color and luminosity that are consistent in amplitude, but not in direction, with those required to produce the gradient we have detected.
+This rotation gradient suggests that main sequence evolution produces offsets in color--magnitude space that are significantly more orthogonal to the zero-age main sequence than models currently predict, and may provide new tests for both stellar evolution and gyrochronology models.
 \end{abstract}
@@
 \centering
-\includegraphics[width=3.6in]{../figures/cmd}
+%\includegraphics[width=3.6in]{../figures/cmd}
+\includegraphics[width=3.6in]{fig1}
 \caption{
@@
 \centering
-\includegraphics[width=3.5in]{../figures/rot_dist_0}
-\includegraphics[width=3.5in]{../figures/rot_dist_350}
-\includegraphics[width=3.5in]{../figures/rot_dist_525}
-\includegraphics[width=3.5in]{../figures/rot_dist_600}
-\includegraphics[width=3.5in]{../figures/rot_dist_700}
-\includegraphics[width=3.5in]{../figures/rot_dist_900}
+%\includegraphics[width=3.5in]{../figures/rot_dist_0}
+%\includegraphics[width=3.5in]{../figures/rot_dist_350}
+%\includegraphics[width=3.5in]{../figures/rot_dist_525}
+%\includegraphics[width=3.5in]{../figures/rot_dist_600}
+%\includegraphics[width=3.5in]{../figures/rot_dist_700}
+%\includegraphics[width=3.5in]{../figures/rot_dist_900}
+\includegraphics[width=3.5in]{fig2a}
+\includegraphics[width=3.5in]{fig2b}
+\includegraphics[width=3.5in]{fig2c}
+\includegraphics[width=3.5in]{fig2d}
+\includegraphics[width=3.5in]{fig2e}
+\includegraphics[width=3.5in]{fig2f}
 \caption{Color--period diagrams for our sample of likely main sequence stars, divided into six bins of distance.
@@
 \centering
-\includegraphics[width=3.5in]{../figures/delta_per_2gauss}
-\includegraphics[width=3.5in]{../figures/delta_per_2gauss_Z}
+%\includegraphics[width=3.5in]{../figures/delta_per_2gauss}
+%\includegraphics[width=3.5in]{../figures/delta_per_2gauss_Z}
+\includegraphics[width=3.5in]{fig3a}
+\includegraphics[width=3.5in]{fig3b}
 \caption{Left: Distribution of the log rotation periods after a 600 Myr gyrochrone was subtracted, using the same distance bins shown in Figure \ref{fig:color_period} (faint lines).
@@
 \centering
-\includegraphics[width=3.4in]{../figures/dist_Z}
-\includegraphics[width=3.5in]{../figures/delta_per_subZ}
+%\includegraphics[width=3.4in]{../figures/dist_Z}
+%\includegraphics[width=3.5in]{../figures/delta_per_subZ}
+\includegraphics[width=3.4in]{fig4a}
+\includegraphics[width=3.5in]{fig4b}
 \caption{
@@
 \centering
-\includegraphics[width=7in]{../figures/cmd_zoom}
+%\includegraphics[width=7in]{../figures/cmd_zoom}
+\includegraphics[width=7in]{fig5}
 \caption{
@@
 The stellar evolution model shows that a star should evolve essentially {\it along} the main sequence track, as in the left panel of Figure \ref{fig:cmd_zoom}.
-The tension with our observed sample is highlighted in the right panel of Figure \ref{fig:cmd_zoom}, where the predicted colors for old stars get bluer, while our slice through the observed rotation period sample shows older stars being redder.
+The tension with our observed sample is highlighted in the right panel of Figure \ref{fig:cmd_zoom}, where the model predictions show stars becoming bluer in $G_{BP}-G_{RP}$ color as they age, while our slice through the observed rotation period sample suggests the older stars are instead redder.
+
+%
+We note that correlations in stellar rotation, age, and color have previously been reported for young clusters \citep[e.g.][]{stauffer2003,covey2016}.
+However, these detections do not provide a straightforward explanation for the offset we detect in the {\em Gaia} CMD for field stars.
+\citet{stauffer2003} detected an offset similar to the gradient we have observed for K dwarfs in the Pleiades and Praesepe open clusters, where younger, more rapidly rotating Pleiades members appear bluer/fainter in the ($B-V$, $V$) CMD than their older, more slowly rotating analogs in Praesepe.
+\citet{stauffer2003} suggest this offset reflects the influence of hot and cool star spots on a star's colors, as supported by the wavelength dependence of the offset.
+The color offset disappears in the ($V-I$, $V$) CMD and reverses in the ($V-K$, $V$) CMD, where the faster rotating Pleiades stars appear redder/brighter than their Praesepe counterparts.
+\citet{covey2016} extended this analysis to demonstrate that a Pleiades member's rotation period correlates with its relative position on the $V-K$ vs.
+$V$ cluster sequence, further supporting a picture in which rapidly rotating, magnetically active stars display different photospheric properties than their slower rotating, less magnetically active brethren.
+The CMD gradient reported here are not easily explained as photospheric signatures of rotationally induced magnetic activity, however, because the gradients are most prominent among the most slowly rotating stars in the sample.
+Indeed, we see no significant changes in the color or magnitude of stars with rotation periods shorter than $\sim$20 days; it is only among the more slowly rotating stars where color and luminosity gradients are significantly detected.
+These slowly rotating stars presumably possess the lowest levels of magnetic activity, and thus have smaller, less prominent starspots to drive changes in the star's bulk photospheric properties.
 
 This tension in the CMD evolution is compounded when we consider stellar metallicity, which we have not varied in our MIST model realization.
-Younger, presumably more metal rich stars should have redder optical colors on the main sequence, while older metal poor stars should be more blue.
-Instead we find that older, slower rotating stars are redder and brighter than their rapidly rotating counterparts.
-While we do not have independent metallicity constraints for the \Kepler stars in this sample, Figure \ref{fig:cmd_zoom} shows that age may have as big of a factor as metallicity in the precise CMD location of main sequence stars.
+Younger, presumably more metal rich stars should have redder optical colors on the main sequence, while older, more metal poor stars should be bluer.
+Instead we find that oldest, more slowly rotating stars are redder and brighter than their rapidly rotating counterparts.
+While we do not have independent metallicity constraints for the \Kepler stars in this sample, Figure \ref{fig:cmd_zoom} shows that age may be justa s important as metallicity in determining the precise CMD location for main main sequence stars.
 Likewise, stellar evolution models do not appear to reproduce the CMD positions for main sequence, low-mass stars of all ages.
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\vspace{0.5in}
 \section{Discussion}
@@
 \acknowledgments
+The authors wish to thank Garrett Somers, Stephanie Douglas, Aaron Dotter, and Jennifer van Saders for their helpful suggestions and discussions during the preparation of this manuscript.
 
@@
 
-\bibliography{/Users/james/Dropbox/references}
+%\bibliography{/Users/james/Dropbox/references}
+
+\begin{thebibliography}{}
+\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
+
+\bibitem[{{Aigrain} {et~al.}(2015){Aigrain}, {Llama}, {Ceillier}, {Chagas},
+  {Davenport}, {Garc{\'{\i}}a}, {Hay}, {Lanza}, {McQuillan}, {Mazeh}, {de
+  Medeiros}, {Nielsen}, \& {Reinhold}}]{aigrain2015}
+{Aigrain}, S., {Llama}, J., {Ceillier}, T., {et~al.} 2015, \mnras, 450, 3211
+
+\bibitem[{{Anderson} {et~al.}(2017){Anderson}, {Hogg}, {Leistedt},
+  {Price-Whelan}, \& {Bovy}}]{anderson2017}
+{Anderson}, L., {Hogg}, D.~W., {Leistedt}, B., {Price-Whelan}, A.~M., \&
+  {Bovy}, J. 2017, ArXiv e-prints, arXiv:1706.05055
+
+\bibitem[{{Angus} {et~al.}(2015){Angus}, {Aigrain}, {Foreman-Mackey}, \&
+  {McQuillan}}]{angus2015}
+{Angus}, R., {Aigrain}, S., {Foreman-Mackey}, D., \& {McQuillan}, A. 2015,
+  \mnras, 450, 1787
+
+\bibitem[{{Astropy Collaboration} {et~al.}(2013){Astropy Collaboration},
+  {Robitaille}, {Tollerud}, {Greenfield}, {Droettboom}, {Bray}, {Aldcroft},
+  {Davis}, {Ginsburg}, {Price-Whelan}, {Kerzendorf}, {Conley}, {Crighton},
+  {Barbary}, {Muna}, {Ferguson}, {Grollier}, {Parikh}, {Nair}, {Unther},
+  {Deil}, {Woillez}, {Conseil}, {Kramer}, {Turner}, {Singer}, {Fox}, {Weaver},
+  {Zabalza}, {Edwards}, {Azalee Bostroem}, {Burke}, {Casey}, {Crawford},
+  {Dencheva}, {Ely}, {Jenness}, {Labrie}, {Lim}, {Pierfederici}, {Pontzen},
+  {Ptak}, {Refsdal}, {Servillat}, \& {Streicher}}]{astropy}
+{Astropy Collaboration}, {Robitaille}, T.~P., {Tollerud}, E.~J., {et~al.} 2013,
+  \aap, 558, A33
+
+\bibitem[{{Bailer-Jones} {et~al.}(2018){Bailer-Jones}, {Rybizki}, {Fouesneau},
+  {Mantelet}, \& {Andrae}}]{bailer-jones2018}
+{Bailer-Jones}, C.~A.~L., {Rybizki}, J., {Fouesneau}, M., {Mantelet}, G., \&
+  {Andrae}, R. 2018, ArXiv e-prints, arXiv:1804.10121
+
+\bibitem[{{Barnes}(2010)}]{barnes2010}
+{Barnes}, S.~A. 2010, \apj, 722, 222
+
+\bibitem[{{Borucki} {et~al.}(2010){Borucki}, {Koch}, {Basri}, {Batalha},
+  {Brown}, {Caldwell}, {Caldwell}, {Christensen-Dalsgaard}, {Cochran},
+  {DeVore}, {Dunham}, {Dupree}, {Gautier}, {Geary}, {Gilliland}, {Gould},
+  {Howell}, {Jenkins}, {Kondo}, {Latham}, {Marcy}, {Meibom}, {Kjeldsen},
+  {Lissauer}, {Monet}, {Morrison}, {Sasselov}, {Tarter}, {Boss}, {Brownlee},
+  {Owen}, {Buzasi}, {Charbonneau}, {Doyle}, {Fortney}, {Ford}, {Holman},
+  {Seager}, {Steffen}, {Welsh}, {Rowe}, {Anderson}, {Buchhave}, {Ciardi},
+  {Walkowicz}, {Sherry}, {Horch}, {Isaacson}, {Everett}, {Fischer}, {Torres},
+  {Johnson}, {Endl}, {MacQueen}, {Bryson}, {Dotson}, {Haas}, {Kolodziejczak},
+  {Van Cleve}, {Chandrasekaran}, {Twicken}, {Quintana}, {Clarke}, {Allen},
+  {Li}, {Wu}, {Tenenbaum}, {Verner}, {Bruhweiler}, {Barnes}, \&
+  {Prsa}}]{borucki2010}
+{Borucki}, W.~J., {Koch}, D., {Basri}, G., {et~al.} 2010, Science, 327, 977
+
+\bibitem[{{Choi} {et~al.}(2016){Choi}, {Dotter}, {Conroy}, {Cantiello},
+  {Paxton}, \& {Johnson}}]{MIST}
+{Choi}, J., {Dotter}, A., {Conroy}, C., {et~al.} 2016, \apj, 823, 102
+
+\bibitem[{{Choi} {et~al.}(2015){Choi}, {Dalcanton}, {Williams}, {Weisz},
+  {Skillman}, {Fouesneau}, \& {Dolphin}}]{choi2015}
+{Choi}, Y., {Dalcanton}, J.~J., {Williams}, B.~F., {et~al.} 2015, \apj, 810, 9
+
+\bibitem[{{Covey} {et~al.}(2016){Covey}, {Ag{\"u}eros}, {Law}, {Liu}, {Ahmadi},
+  {Laher}, {Levitan}, {Sesar}, \& {Surace}}]{covey2016}
+{Covey}, K.~R., {Ag{\"u}eros}, M.~A., {Law}, N.~M., {et~al.} 2016, \apj, 822,
+  81
+
+\bibitem[{{Davenport}(2017)}]{davenport2017}
+{Davenport}, J.~R.~A. 2017, \apj, 835, 16
+
+\bibitem[{{do Nascimento} {et~al.}(2012){do Nascimento}, {da Costa}, \&
+  {Castro}}]{donascimento2012}
+{do Nascimento}, J.-D., {da Costa}, J.~S., \& {Castro}, M. 2012, \aap, 548, L1
+
+\bibitem[{{Gaia Collaboration} {et~al.}(2018){Gaia Collaboration}, {Brown},
+  {Vallenari}, {Prusti}, {de Bruijne}, {Babusiaux}, \&
+  {Bailer-Jones}}]{gaia_dr2}
+{Gaia Collaboration}, {Brown}, A.~G.~A., {Vallenari}, A., {et~al.} 2018, ArXiv
+  e-prints, arXiv:1804.09365
+
+\bibitem[{{Gaia Collaboration} {et~al.}(2016){Gaia Collaboration}, {Prusti},
+  {de Bruijne}, {Brown}, {Vallenari}, {Babusiaux}, {Bailer-Jones}, {Bastian},
+  {Biermann}, {Evans}, \& et~al.}]{gaia}
+{Gaia Collaboration}, {Prusti}, T., {de Bruijne}, J.~H.~J., {et~al.} 2016,
+  \aap, 595, A1
+
+\bibitem[{{Gilmore} \& {Reid}(1983)}]{gilmore1983}
+{Gilmore}, G., \& {Reid}, N. 1983, \mnras, 202, 1025
+
+\bibitem[{{Howell} {et~al.}(2014){Howell}, {Sobeck}, {Haas}, {Still},
+  {Barclay}, {Mullally}, {Troeltzsch}, {Aigrain}, {Bryson}, {Caldwell},
+  {Chaplin}, {Cochran}, {Huber}, {Marcy}, {Miglio}, {Najita}, {Smith},
+  {Twicken}, \& {Fortney}}]{howell2014}
+{Howell}, S.~B., {Sobeck}, C., {Haas}, M., {et~al.} 2014, \pasp, 126, 398
+
+\bibitem[{Hunter(2007)}]{matplotlib}
+Hunter, J.~D. 2007, Computing In Science \& Engineering, 9, 90
+
+\bibitem[{{Jao} {et~al.}(2018){Jao}, {Henry}, {Gies}, \& {Hambly}}]{jao2018}
+{Jao}, W.-C., {Henry}, T.~J., {Gies}, D.~R., \& {Hambly}, N.~C. 2018, ArXiv
+  e-prints, arXiv:1806.07792
+
+\bibitem[{Jones {et~al.}(2001--)Jones, Oliphant, Peterson, {et~al.}}]{scipy}
+Jones, E., Oliphant, T., Peterson, P., {et~al.} 2001--, {SciPy}: Open source
+  scientific tools for {Python}, , , [Online; accessed <today>]
+
+\bibitem[{{Lanzafame} {et~al.}(2018){Lanzafame}, {Distefano}, {Messina},
+  {Pagano}, {Lanza}, {Eyer}, {Guy}, {Rimoldini}, {Lecoeur-Taibi}, {Holl}, {de
+  Fombelle}, {Nienartowicz}, {Marchal}, \& {Mowlavi}}]{lanzafame2018}
+{Lanzafame}, A.~C., {Distefano}, E., {Messina}, S., {et~al.} 2018, ArXiv
+  e-prints, arXiv:1805.00421
+
+\bibitem[{{Lewis} {et~al.}(2015){Lewis}, {Dolphin}, {Dalcanton}, {Weisz},
+  {Williams}, {Bell}, {Seth}, {Simones}, {Skillman}, {Choi}, {Fouesneau},
+  {Guhathakurta}, {Johnson}, {Kalirai}, {Leroy}, {Monachesi}, {Rix}, \&
+  {Schruba}}]{lewis2015}
+{Lewis}, A.~R., {Dolphin}, A.~E., {Dalcanton}, J.~J., {et~al.} 2015, \apj, 805,
+  183
+
+\bibitem[{{Lurie} {et~al.}(2017){Lurie}, {Vyhmeister}, {Hawley}, {Adilia},
+  {Chen}, {Davenport}, {Juri{\'c}}, {Puig-Holzman}, \&
+  {Weisenburger}}]{lurie2017}
+{Lurie}, J.~C., {Vyhmeister}, K., {Hawley}, S.~L., {et~al.} 2017, \aj, 154, 250
+
+\bibitem[{{Ma} {et~al.}(2017){Ma}, {Hopkins}, {Wetzel}, {Kirby},
+  {Angl{\'e}s-Alc{\'a}zar}, {Faucher-Gigu{\`e}re}, {Kere{\v s}}, \&
+  {Quataert}}]{ma2017}
+{Ma}, X., {Hopkins}, P.~F., {Wetzel}, A.~R., {et~al.} 2017, \mnras, 467, 2430
+
+\bibitem[{{Matt} {et~al.}(2015){Matt}, {Brun}, {Baraffe}, {Bouvier}, \&
+  {Chabrier}}]{matt2015}
+{Matt}, S.~P., {Brun}, A.~S., {Baraffe}, I., {Bouvier}, J., \& {Chabrier}, G.
+  2015, \apjl, 799, L23
+
+\bibitem[{McKinney(2010)}]{pandas}
+McKinney, W. 2010, in Proceedings of the 9th Python in Science Conference, ed.
+  S.~van~der Walt \& J.~Millman, 51 -- 56
+
+\bibitem[{{McQuillan} {et~al.}(2013){McQuillan}, {Aigrain}, \&
+  {Mazeh}}]{mcquillan2013}
+{McQuillan}, A., {Aigrain}, S., \& {Mazeh}, T. 2013, \mnras, 432, 1203
+
+\bibitem[{{McQuillan} {et~al.}(2014){McQuillan}, {Mazeh}, \&
+  {Aigrain}}]{mcquillan2014}
+{McQuillan}, A., {Mazeh}, T., \& {Aigrain}, S. 2014, \apjs, 211, 24
+
+\bibitem[{{Meibom} {et~al.}(2009){Meibom}, {Mathieu}, \&
+  {Stassun}}]{meibom2009}
+{Meibom}, S., {Mathieu}, R.~D., \& {Stassun}, K.~G. 2009, \apj, 695, 679
+
+\bibitem[{{Meibom} {et~al.}(2011){Meibom}, {Barnes}, {Latham}, {Batalha},
+  {Borucki}, {Koch}, {Basri}, {Walkowicz}, {Janes}, {Jenkins}, {Van Cleve},
+  {Haas}, {Bryson}, {Dupree}, {Furesz}, {Szentgyorgyi}, {Buchhave}, {Clarke},
+  {Twicken}, \& {Quintana}}]{meibom2011}
+{Meibom}, S., {Barnes}, S.~A., {Latham}, D.~W., {et~al.} 2011, \apjl, 733, L9
+
+\bibitem[{Oliphant(2007)}]{numpy}
+Oliphant, T.~E. 2007, Computing in Science Engineering, 9, 10
+
+\bibitem[{P\'erez \& Granger(2007)}]{ipython}
+P\'erez, F., \& Granger, B.~E. 2007, Computing in Science and Engineering, 9,
+  21
+
+\bibitem[{{Sandquist} {et~al.}(2016){Sandquist}, {Jessen-Hansen}, {Shetrone},
+  {Brogaard}, {Meibom}, {Leitner}, {Stello}, {Bruntt}, {Antoci}, {Orosz},
+  {Grundahl}, \& {Frandsen}}]{sandquist2016}
+{Sandquist}, E.~L., {Jessen-Hansen}, J., {Shetrone}, M.~D., {et~al.} 2016,
+  \apj, 831, 11
+
+\bibitem[{{Skumanich}(1972)}]{skumanich1972}
+{Skumanich}, A. 1972, \apj, 171, 565
+
+\bibitem[{{Stauffer} {et~al.}(2003){Stauffer}, {Jones}, {Backman}, {Hartmann},
+  {Barrado y Navascu{\'e}s}, {Pinsonneault}, {Terndrup}, \&
+  {Muench}}]{stauffer2003}
+{Stauffer}, J.~R., {Jones}, B.~F., {Backman}, D., {et~al.} 2003, \aj, 126, 833
+
+\bibitem[{{van Saders} {et~al.}(2016){van Saders}, {Ceillier}, {Metcalfe},
+  {Silva Aguirre}, {Pinsonneault}, {Garc{\'{\i}}a}, {Mathur}, \&
+  {Davies}}]{van-saders2016}
+{van Saders}, J.~L., {Ceillier}, T., {Metcalfe}, T.~S., {et~al.} 2016, \nat,
+  529, 181
+
+\bibitem[{{van Saders} \& {Pinsonneault}(2013)}]{van-saders2013}
+{van Saders}, J.~L., \& {Pinsonneault}, M.~H. 2013, \apj, 776, 67
+
+\bibitem[{{Williams} {et~al.}(2017){Williams}, {Dolphin}, {Dalcanton}, {Weisz},
+  {Bell}, {Lewis}, {Rosenfield}, {Choi}, {Skillman}, \&
+  {Monachesi}}]{williams2017}
+{Williams}, B.~F., {Dolphin}, A.~E., {Dalcanton}, J.~J., {et~al.} 2017, \apj,
+  846, 145
+
+\bibitem[{{Xiang} {et~al.}(2017){Xiang}, {Liu}, {Shi}, {Yuan}, {Huang}, {Chen},
+  {Wang}, {Tian}, {Wu}, {Yang}, {Zhang}, {Huo}, \& {Ren}}]{xiang2017}
+{Xiang}, M., {Liu}, X., {Shi}, J., {et~al.} 2017, \apjs, 232, 2
+
+\end{thebibliography}
 

### 2018-10-11 paper/ms.tex
responded to referee

--- 2018-07-20 paper/ms.tex
@@
-\documentclass[preprint2]{aastex62}
+\documentclass[trackchanges,preprint2]{aastex62}
 
@@
 \correspondingauthor{James. R. A. Davenport}
-\email{James.Davenport@wwu.edu}
+\email{jrad@uw.edu}
 
@@
 The rotation period bimodality, originally discovered by \citet{mcquillan2013}, is clearly recovered for stars out to 525pc, but is not detectable at further distances. 
-This bimodality correlates with Galactic height as well, dropping strongly for stars above $Z>90$ pc.
-We also find a significant width in the stellar main sequence of $\Delta M_G\sim$0.25 mag, as well as a coherent gradient of increasing rotation periods diagonal to the main sequence.
-We interpret this to be the simultaneous change in luminosity and loss of angular momentum as stars evolve along the main sequence.
-The diagonal offset from the main sequence is unexpected, however, and may provide a new independent test of both stellar evolution and gyrochronlogy models.
+\replaced{This bimodality correlates with Galactic height as well, dropping strongly for stars above $Z>90$ pc.}{We find the bimodality is correlated with height above the Galacitc plane, with the ratio of rapidly rotating younger stars dropping strongly for stars above $Z>90$ pc.}
+We also find a significant width in the stellar main sequence of $\Delta M_G\sim$0.25 mag, as well as a coherent gradient of increasing rotation periods orthogonal to the main sequence.
+We interpret this as a signature of stellar angular momentum loss over time, implying a corresponding diagonal age gradient across the main sequence.
+Stellar evolution models predict changes in color and luminosity that are consistent in amplitude, but not in direction, with those required to produce the gradient we have detected.
+This rotation gradient \replaced{suggests}{may indicate} that main sequence evolution produces offsets in color--magnitude space that are significantly more orthogonal to the zero-age main sequence than models currently predict, and may provide new tests for both stellar evolution and gyrochronology models.
 \end{abstract}
@@
 \section{Introduction}
+\label{sec:intro}
+
 Over the past decade the study of stellar rotation has transformed from a niche observation limited to a handful of nearby clusters and few thousand bright stars, to an area of great  interest for field stars with a rapidly growing sample size.
-The \Kepler mission \citep{borucki2010}, through its unmatched combination of photometric precision and years-long observation baseline, has produced the largest precision sample of rotating stars to date -- more than 30,000 sources to date \citep{mcquillan2014}.
+The \Kepler mission \citep{borucki2010}, through its unmatched combination of photometric precision and years-long observation baseline, has produced the largest precision sample of rotating stars to date -- more than 30,000 sources to date \citep{nielsen2013,reinhold2013,mcquillan2014}.
 Although the light curves available from the {\em Gaia} mission are not as densely sampled in time as \Kepler \citep{gaia}, they have yielded over 147,000 candidate rotation periods as of Data Release 2 \citep{lanzafame2018}.
@@
 \section{The \Kepler--Gaia Data}
+\label{sec:data}
 
@@
 The sample from \citet{mcquillan2014} provides rotation periods for more 34,030 stars, measured using the Auto-Correlation Function (ACF).
+\added{Catalogs of \Kepler rotation periods using other methods are available, such as \citet{reinhold2013} and \citet{nielsen2013} that use the Lomb-Scargle Periodogram.}
 While the ACF does not recover periods with as much precision as methods such as the Lomb-Scargle Periodogram, it is more robust to detecting the true period as opposed to an alias, and more complete for batch analysis of all stars \citep[e.g.
@@
 \section{Selecting Main Sequence Stars}
+\label{sec:MS}
 
@@
 We do not explore the binary population in any detail here, but this sample could provide useful insight into the tidal evolution of binary stars, and are good targets for radial velocity follow-up to characterize binary system properties.
-We also note a small number of systems above the even the equal-mass binary main sequence track, which could be due to unresolved triple star systems.
+We also note a small number of systems \added{are present in the {\em Gaia} data} above the even the equal-mass binary main sequence track, which could be due to unresolved triple star systems.
 
@@
 Single, main sequence stars were selected in a region spanning 0.1 mag fainter and 0.4 mag brighter than the MIST isochrone, resulting in a final sample of 16,248 stars for analysis of their rotation period distributions.
+\added{
+While this shifted isochrone cut eliminates the clear binary systems, lower mass ratio binaries may still be included in our sample.
+Without a census for the total binary population in the \Kepler field, however, the impact of contaminating binary stars on our analysis is unknown.
+}
 
@@
 \section{Tracing the Period Bimodality}
+\label{sec:bimodal}
 
@@
 
+% added in revised version
+\begin{figure}[!ht]
+\centering
+%\includegraphics[width=3.5in]{../figures/B_V_rot}
+\includegraphics[width=3.5in]{fig3}
+\caption{Color--period diagram for the nearest bin shown in Fig.
+\ref{fig:color_period}, transformed from {\em Gaia} colors into $B-V$ using the isochrone as described in the text.
+As in \citet{davenport2017}, a 600 Myr gyrochrone from \citet{meibom2009} traces the rotation period bimodality (red line).}
+\label{fig:BV_gyro}
+\end{figure}
+
 \begin{figure*}[!ht]
 \centering
-\includegraphics[width=3.5in]{../figures/delta_per_2gauss}
-\includegraphics[width=3.5in]{../figures/delta_per_2gauss_Z}
+%\includegraphics[width=3.5in]{../figures/delta_per_2gauss}
+%\includegraphics[width=3.5in]{../figures/delta_per_2gauss_Z}
+\includegraphics[width=3.5in]{fig4a}
+\includegraphics[width=3.5in]{fig4b}
 \caption{Left: Distribution of the log rotation periods after a 600 Myr gyrochrone was subtracted, using the same distance bins shown in Figure \ref{fig:color_period} (faint lines).
@@
 
-To better illustrate the evolution of rotation periods for all the stars between the distance bins, in Figure \ref{fig:per_hist} we follow \citet{davenport2017} and subtract the rotation period of a 600 Myr gyrochrone.
+To better illustrate the evolution of rotation periods for all the stars between the distance bins, \deleted{in Figure \ref{fig:per_hist}} we \added{will} follow \citet{davenport2017} and subtract the rotation period of a 600 Myr gyrochrone.
 As no published gyrochronology model yet exists that has been tuned to the {\em Gaia} photometric colors, we adopt the same gyrochronology model of Eqn.
 2 from \citet{meibom2009} used by \citet{davenport2017} to approximately trace the rotation period gap at 600 Myr as a function of $B-V$ color.
-We convert stars from the observed {\em Gaia} $G_{BP}-G_{RP}$ color to $B-V$ using the same 1 Gyr MIST isochrone model used to define the main sequence in Figure \ref{fig:cmd} above.
+We convert stars from the observed {\em Gaia} $G_{BP}-G_{RP}$ color to $B-V$ using the same 1 Gyr MIST isochrone model used to define the main sequence in Figure \ref{fig:cmd} above, \added{shown in Figure \ref{fig:BV_gyro}.}
 
@@
 
-To further understanding the spatial extent of this age-related feature, in Figure \ref{fig:dZ} we break the lowest height stars ($Z < 100$ pc) in to two roughly even samples, split as a function of their projected distance.
+To further understand\deleted{ing} the spatial extent of this age-related feature, in Figure \ref{fig:dZ} we break the lowest height stars ($Z < 100$ pc) in to two roughly even samples, split as a function of their projected distance.
 Note we have also repeated this exercise for stars in higher ranges of $Z$, and find the decline of the rapid rotators is again uniform between subsamples of varying projected distance.
@@
 Instead, we believe this feature is characteristic of the age--$Z$ dynamical correlation observed within the solar neighborhood of our Galaxy.
+\added{We also investigated the total proper motions of the fast (young) versus the slow-rotating (old) populations of stars with our sample, as was done in \citet{mcquillan2013} and \citet{davenport2017}.
+As expected, the two populations systematically show different total proper motion amplitudes at all distances, further supporting the age--$Z$ correlation we have observed.
+}
 
@@
 \centering
-\includegraphics[width=3.4in]{../figures/dist_Z}
-\includegraphics[width=3.5in]{../figures/delta_per_subZ}
+%\includegraphics[width=3.4in]{../figures/dist_Z}
+%\includegraphics[width=3.5in]{../figures/delta_per_subZ}
+\includegraphics[width=3.4in]{fig5a}
+\includegraphics[width=3.5in]{fig5b}
 \caption{
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\section{Recalibrating Stellar Evolution Models}
+\section{An Unexplained Feature in the Color--Magnitude--Period Diagram}
+\label{sec:mystery}
 
@@
 \centering
-\includegraphics[width=7in]{../figures/cmd_zoom}
+%\includegraphics[width=7in]{../figures/cmd_zoom}
+\includegraphics[width=7in]{fig6}
 \caption{
@@
 The stellar evolution model shows that a star should evolve essentially {\it along} the main sequence track, as in the left panel of Figure \ref{fig:cmd_zoom}.
-The tension with our observed sample is highlighted in the right panel of Figure \ref{fig:cmd_zoom}, where the predicted colors for old stars get bluer, while our slice through the observed rotation period sample shows older stars being redder.
+The tension with our observed sample is highlighted in the right panel of Figure \ref{fig:cmd_zoom}, where the model predictions show stars becoming bluer in $G_{BP}-G_{RP}$ color as they age, while our slice through the observed rotation period sample suggests the older stars are instead redder.
+
+We note that correlations in stellar rotation, age, and color have previously been reported for young clusters \citep[e.g.][]{stauffer2003,covey2016}.
+However, these detections do not provide a straightforward explanation for the offset we detect in the {\em Gaia} CMD for field stars.
+\citet{stauffer2003} detected an offset similar to the gradient we have observed for K dwarfs in the Pleiades and Praesepe open clusters, where younger, more rapidly rotating Pleiades members appear bluer/fainter in the ($B-V$, $V$) CMD than their older, more slowly rotating analogs in Praesepe.
+\citet{stauffer2003} suggest this offset reflects the influence of hot and cool star spots on a star's colors, as supported by the wavelength dependence of the offset.
+The color offset disappears in the ($V-I$, $V$) CMD and reverses in the ($V-K$, $V$) CMD, where the faster rotating Pleiades stars appear redder/brighter than their Praesepe counterparts.
+\citet{covey2016} extended this analysis to demonstrate that a Pleiades member's rotation period correlates with its relative position on the $V-K$ vs.
+$V$ cluster sequence, further supporting a picture in which rapidly rotating, magnetically active stars display different photospheric properties than their slower rotating, less magnetically active brethren.
+The CMD gradient reported here are not easily explained as photospheric signatures of rotationally induced magnetic activity, however, because the gradients are most prominent among the most slowly rotating stars in the sample.
+Indeed, we see no significant changes in the color or magnitude of stars with rotation periods shorter than $\sim$20 days; it is only among the more slowly rotating stars where color and luminosity gradients are significantly detected.
+These slowly rotating stars presumably possess the lowest levels of magnetic activity, and thus have smaller, less prominent starspots to drive changes in the star's bulk photospheric properties.
 
 This tension in the CMD evolution is compounded when we consider stellar metallicity, which we have not varied in our MIST model realization.
-Younger, presumably more metal rich stars should have redder optical colors on the main sequence, while older metal poor stars should be more blue.
-Instead we find that older, slower rotating stars are redder and brighter than their rapidly rotating counterparts.
-While we do not have independent metallicity constraints for the \Kepler stars in this sample, Figure \ref{fig:cmd_zoom} shows that age may have as big of a factor as metallicity in the precise CMD location of main sequence stars.
+Younger, presumably more metal rich stars should have redder optical colors on the main sequence, while older, more metal poor stars should be bluer.
+Instead we find that oldest, more slowly rotating stars are redder and brighter than their rapidly rotating counterparts.
+While we do not have independent metallicity constraints for the \Kepler stars in this sample, Figure \ref{fig:cmd_zoom} shows that age may be \replaced{justa s}{just as} important as metallicity in determining the precise CMD location for main main sequence stars.
 Likewise, stellar evolution models do not appear to reproduce the CMD positions for main sequence, low-mass stars of all ages.
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\vspace{0.5in}
 \section{Discussion}
+\label{sec:discussion}
 
 Using a sample of 16,248 single main sequence stars with measured rotation periods from \Kepler and parallaxes from {\em Gaia} DR2, we have begun to explore the spatial distribution of stellar ages near the Sun using gyrochronology.
-The bimodality in rotation periods first reported by \citet{mcquillan2013} appear to be constrained to low Galactic scaleheights, rather than in an obvious bubble centered around the Sun.
-Since height above the Galactic plane is assumed to be related to age, this is consistent with the rotation period bimodality being a direct tracer of the star formation history, and indicates a burst of star formation within the past $\sim$600 Myr within the Galactic disk.
+The bimodality in rotation periods first reported by \citet{mcquillan2013} appear to be constrained to low Galactic scale-heights, rather than in an obvious bubble centered around the Sun.
+Since height above the Galactic plane is assumed to be related to age, this is consistent with the rotation period bimodality being a direct tracer of the star formation history, and indicates a burst of star formation within the past $\sim$600 Myr.
+\deleted{ within the Galactic disk.
+}
+
+\added{Independently verifying the star formation history implied by the rotation period bimodality, particularly over the past 1 Gyr, is difficult.
+White dwarf ``cosmochronology'' is sensitive to star formation over much longer timescales \citep{tremblay2014}.
+Color--Magnitude diagram inversion has found some variation in star formation in the solar neighbhorhood within the past $\sim$1 Gyr \citep{hernandez2000,cignoni2006}.
+However, these studies have not yet tapped the full potential of {\em Gaia} to study short timescale variations or spatial distributions in star formation history \citep{bertelli1999,bernard2018}.
+Within the past 600 Myr, the Sun has likely lapped the entire spiral pattern of the Milky Way, passing through multiple spiral arms \citep{svensmark2006}.
+Whether the young stars in the rotation period bimodality represents a Galaxy-wide recent burst of star formation remains unclear.
+}
 
@@
 For example, comparing the rotation period distribution for binary versus single main sequence stars may provide constraints on their dynamical histories.
+\added{
+Interestingly, in a preliminary exploration of the equal-mass binary stars from this \Kepler-{\em Gaia} sample, we find the rotation period bimodality is visible for nearby systems.
+}
+
+\added{
+In our investigation of the {\em Gaia}--\Kepler color--magnitude--rotation period diagram, we have discovered a new feature: a diagonal gradient of decreasing rotation periods across the main sequence, suggesting that older (slower rotating) stars are brighter and redder than their younger counterparts.
+This structure is wholly unexpected, both from observation and theory.
+We have illustrated the theoretical expectation in Fig.
+\ref{fig:cmd_zoom} using isochrone models, that stars should evolve roughly {\it along} the main sequence during their lifetimes, rather than orthogonally as our data suggest.
+Observations of the main sequences from open and globular clusters with {\em Gaia} have not revealed any similar diagonal evolution, though a detailed comparison for low-mass stars has not yet been done \citep{babusiaux2018}.
+The period gradient we have discovered represents either an intriguing new detail of stellar evolution, or a bias in the \Kepler or {\em Gaia} data that is not presently understood.
+}
+
+\added{
+Note that our sample is contaminated by some unknown number of low mass ratio binary stars, which were not perfectly removed using our isochrone filtering in \S\ref{sec:MS}.
+These low mass ratio systems systematically lie slightly brighter and redder than the single star main sequence, exactly where the rotation period gradient feature appears strongest.
+However, the period gradient would put very strong constraints on the nature and evolution of these binary stars.
+For example, to generate the shift in the average rotation period seen in our Fig.
+\ref{fig:cmd_zoom}, binaries at this mass ratio must all have orbits short enough to tidally synchronize the rotation periods we observe with \Kepler.
+This would suggest that A) virtually no long orbital period binaries are present in the population, and B) their orbits are synchronizing at 30-40 days, rather than the 10-20 days expected \citep[e.g.][]{lurie2017}.
+As noted above, the presence of the \Kepler rotation period bimodality for the near equal-mass binaries would further indicate that something unique was occurring dynamically for only these low mass ratio systems.
+While binary stars are a clear candidate due to their complex tidal evolution (D.
+Flemming et al.
+2018 {\em in prep}) and location on the color--magnitude diagram, the mechanism to create the observed rotation period gradient is not clear.
+}
 
@@
 \acknowledgments
+The authors wish to thank Garrett Somers, Stephanie Douglas, Aaron Dotter, and Jennifer van Saders for their helpful suggestions and discussions during the preparation of this manuscript, and the anonymous referee whose comments improved our discussion of the results.
 
